  \noindent We wrote this book during the 1990s, 2000s, and  2010s while
teaching graduate courses
in macro and monetary economics.
 We owe a substantial  debt to
the students in these classes for learning with us. We would
especially like to thank Marco Bassetto,
Victor Chernozhukov,
Riccardo Colacito, Mariacristina De Nardi, William Dupor, William
Fuchs, George Hall, Sagiri Kitao, Hanno Lustig,
 Monika Piazzesi, Navin Kartik,
Martin Schneider,   Yongseok Shin, Christopher
Sleet, Stijn Van Nieuwerburgh, Laura Veldkamp, Neng  Wang, Chao
Wei, Mark Wright, and Sevin Yeltekin for commenting on drafts of earlier editions when they
were graduate students.  In prefaces to earlier editions, we forecast that they would soon be prominent
economists, and we are happy that  has come true.
We also thank
 Isaac Baley, Anmol Bhandari,  Saki Bigio,  Jaroslav Borovicka,  David Evans, Sebastian Graves, Christopher Huckfeldt, Anna Orlik,  Ignacio Presno, Cecilia Parlatore Siritto,  Balint Sz\"oke, and    H\aa k\.on Travoll for helpful comments
 on earlier drafts of this edition.  Each
of these people made substantial suggestions for improving this
book. We expect much from members of this group, as we did from an
earlier group of students that Sargent  (1987b) thanked.


We received useful
 criticisms from Roberto Chang,
Gary Hansen, Jonathan Heathcote, Berthold Herrendorf,  Mark
Huggett, Charles Jones, Narayana Kocherlakota, Dirk Krueger, Per
Krusell, Francesco Lippi, Rodolfo Manuelli, Beatrix Paal, Adina
Popescu, Jonathan Thomas,  Nicola Tosini, and Jesus Fernandez-Villaverde.


Rodolfo Manuelli and Pierre Olivier Weill kindly allowed us to reproduce some of their
exercises. We indicate the exercises that  they donated.
 Some of the exercises in chapters
\use{search1}, \use{ogmodels},  and \use{townsend} are versions of ones in Sargent (1987b).
Fran\c cois Velde provided  substantial help with the \TeX \ and
Unix macros that produced this book.
 Maria Bharwada helped  typeset the first edition.
%We thank P.M.\ Gordon Associates for copyediting.

For providing good environments to work on this book, Ljungqvist
thanks the Stockholm School of Economics and Sargent thanks the
Hoover Institution and the departments of economics at the
University of Chicago, Stanford University, Princeton University, and New York
University.
