% Bibliography macros
%
%\def\refshape{\parshape 2 0truecm 14.765truecm .5truecm
%14.265truecm\noindent} %  old ones




%\def\refshape{\parshape 2 0truecm 10.565truecm .5truecm
%10.565truecm\noindent}

%\def\refshape{\parshape 2 0truecm 12.565truecm .5truecm
%10.565truecm\noindent}
%  the above works pretty well

\def\refshape{\parshape 2 0truecm 12.165truecm .5truecm
10.565truecm\noindent}


\def\book#1#2#3#4{\refshape #1. #2.  {\sl #3}. #4.}



\def\article#1#2#3#4#5{\refshape #1. #2. ``#3'' {\it #4},
#5.}


\def\contribution#1#2#3#4#5#6{\refshape #1. #2. ``#3.'' In #4
(eds.), {\it #5}. #6.}

\def\pcontribution#1#2#3#4#5#6{\refshape #1. #2. ``#3.'' In #4
(eds.), {\it #5}. #6}


\def\ocontribution#1#2#3#4#5#6{\refshape #1. #2. ``#3.'' In #4
(ed.), {\it #5} #6.}


\def\mimeo#1#2#3#4{\refshape #1. #2. ``#3''  Mimeo. #4.}


\def\dissertation#1#2#3#4{\refshape #1. #2. ``#3.'' Ph.D.\
dissertation, #4.}
{\eightpoint


 \article{Abel, Andrew B., N. Gregory Mankiw, Lawrence H. Summers,
 and Richard J. Zeckhauser}{1989}{Assessing Dynamic Efficiency: Theory and
 Evidence.}{Review of Economic Studies}{Vol. 56,  pp. 1--20}




 \article{Abreu, Dilip}{1988}{On the Theory of Infinitely Repeated Games
 with Discounting.}{Econometrica}{Vol. 56, pp. 383--396}


 \article{Abreu, Dilip, David Pearce, and Ennio Stacchetti}{1986}
 {Optimal Cartel Equilibria with Imperfect Monitoring.}
 {Journal of Economic Theory}{Vol. 39, pp. 251--269}


 \article{Abreu, Dilip, David Pearce, and Ennio Stacchetti}{1990}
 {Toward a Theory of Discounted Repeated Games with Imperfect Monitoring.}
 {Econometrica}{Vol. 58(5), pp. 1041--1063}


\mimeo{Acemoglu, Daron}{1997}{Good Jobs versus Bad Jobs: Theory and Some
Evidence.}{CEPR Discussion Paper No. 1588}


\article{Acemoglu, Daron, and Robert Shimer}{1999}{Efficient Unemployment
Insurance.}{Journal of Political Economy}{Vol. 107, pp. 893--928}




\article{Acemoglu, Daron, and Robert Shimer}{2000}
{Productivity gains from unemployment insurance.}
{European Economic
Review}{Vol. 44, pp. 1195--1224}




\book{Adda, Jerome, and Russell W. Cooper}{2003}{Dynamic
Economics: Quantitative Methods and Applications}{Cambridge, Mass.:
MIT Press}


\article{Aghion, Philippe, and Peter Howitt}{1992}{A Model of Growth through
Creative Destruction.}{Econometrica}{Vol. 60, pp. 323--351}


\book{Aghion, Philippe, and Peter Howitt}{1998}{Endogenous Growth Theory}
{Cambridge, Mass.: MIT Press}


 \article{Aiyagari, S. Rao}{1985}{Observational Equivalence of the Overlapping
 Generations and the Discounted Dynamic Programming Frameworks for
 One-Sector Growth.}{Journal of Economic Theory}{Vol. 35(2), pp. 201--221}


 \article{Aiyagari, S. Rao}{1987}{Optimality and Monetary Equilibria in Stationary Overlapping Generations Models with Long Lived Agents.}
 {Journal of Economic Theory}{Vol. 43, pp. 292--313}


 \article{Aiyagari, S. Rao}{1993}{Explaining Financial Market Facts: The Importance of Incomplete Markets and Transaction Costs.}{Quarterly Review}
{Federal Reserve Bank of Minneapolis, Vol. 17(1), pp. 17--31}


 \article{Aiyagari, S. Rao}{1994}{Uninsured Idiosyncratic Risk and Aggregate
 Saving.}{Quarterly Journal of Economics}{Vol. 109(3), pp. 659--684}


 \article{Aiyagari, S. Rao}{1995}{Optimal Capital Income Taxation with Incomplete Markets and Borrowing Constraints.}{Journal of Political Economy}
{Vol. 103(6), pp. 1158--1175}


 \article{Aiyagari, S. Rao, and Mark Gertler}{1991}{Asset Returns with
 Transactions Costs and Uninsured Individual Risk.}{Journal of
 Monetary Economics}{Vol. 27, pp. 311--331}


 \article{Aiyagari, S. Rao, and Ellen R. McGrattan}{1998}
 {The Optimum Quantity of Debt.}{Journal of Monetary Economics}
 {Vol. 42(3), pp. 447--469}


 \article{Aiyagari, S. Rao, and Neil Wallace}{1991}{Existence of Steady
 States with Positive Consumption in the Kiyotaki-Wright Model.}
 {Review of Economic Studies}{Vol. 58(5), pp. 901--916}


\article{Aiyagari, Marcet, Albert, Thomas J. Sargent, and Juha Sepp\"al\"a}
{2002}{Optimal Taxation without State-Contingent Debt.}
{Journal of Political Economy}{Vol. 110(6), pp. 1220--1254}



\article{Albarran, Pedro,  and Orazio Attanasio}{2003}{Limited
Commitment and Crowding out of Private Transfers: Evidence from a
Randomized Experiment}{Economic Journal}{Vol. 113(486), pp.
C77--85}


 \article{Albrecht, James, and Bo Axell}{1984}{An Equilibrium Model
 of Search Unemployment.}{Journal of Political Economy}
 {Vol. 92(5), pp. 824--840}

 \article{Allen, Franklin}{1985}{Repeated principal-agent relationships with lending and borrowing.}
 {Economics Letters}{Vol. 17(1-2), pp. 27-31}



 \article{Altug, Sumru}{1989}{Time-to-Build and Aggregate Fluctuations:
 Some New Evidence.}{International Economic Review}
 {Vol. 30(4), pp. 889--920}


 \book{Altug, Sumru, and Pamela Labadie}{1994}{Dynamic Choice and Asset
 Markets}{San Diego: Academic Press}





\article{Alvarez, Fernando, and Urban J. Jermann}{2000}{Efficiency,
Equilibrium, and Asset Pricing with Risk of
Default}{Econometrica}{Vol. 68(4), pp. 775-797}




\article{Alvarez, Fernando, and Urban J.
Jermann}{2001}{Quantitative Asset Pricing Implications of
Endogenous Solvency Constraints.}{Review of Financial Studies}
{Vol. 14(4), pp.  1117-1151}


\article{Alvarez, Fernando, and Urban J. Jermann}{2004}{Using Asset Prices to Measure
 the Cost  of Business Cycles.}{Journal of Political Economy}{Vol. 112, No. 6, pp. 1223--1256}


 \article{Alvarez, Fernando, Patrick J. Kehoe, and Pablo Andr\'es Neumeyer}{2004}
{The Time Consistency of Optimal Monetary and Fiscal Policies.}{Econometrica}{Vol. 72, pp. 541--567}

 \contribution{Anderson, Evan W., Lars P. Hansen, Ellen R. McGrattan,
 and Thomas J. Sargent}{1996}{Mechanics of Forming and Estimating
 Dynamic Linear Economies}{Hans M. Amman, David A. Kendrick,
 and John Rust}{Handbook of Computational Economics, Vol. 1,
 Handbooks in Economics, Vol. 13}{Amsterdam:
 Elsevier Science, North-Holland, pp. 171--252}

\article{Anderson, E., L. P. Hansen, and T. J. Sargent}{2003} {A
Quartet of Semigroups for Model Specification, Robustness, Prices of
Risk, and Model Detection.} {Journal of the European Economic
Association} {Vol. 1(1), pp. 68--123}



\article{Ang, Andrew, and Monika Piazzesi}{2003}{A No-Arbitrage Vector
Autoregression of Term Structure Dynamics with Macroeonomic and
Latent Variables.}{Journal of Monetary Economics}{Vol. 50(4),
pp. 745--288}


\article{Angeletos, George-Marios}{2002}{Fiscal Policy With Noncontingent Debt And The Optimal Maturity Structure}{Quarterly Journal of Economics}{Vol. 117(3), pp. 1105-1131}



 \book{Apostol, Tom M}{1975}{Mathematical Analysis}{2nd ed.
 Reading, Mass.: Addison-Wesley}


\article{Arrow,
Kenneth J}{1962}{The Economic Implications of
 Learning by Doing.}{Review of Economic Studies}{Vol. 29, pp. 155--173}


\article{Arrow, Kenneth J}{1964}{The Role of Securities in the Optimal
Allocation of Risk-Bearing.}{Review of Economic Studies}{Vol. 31, pp. 91--96}


 \article{\AA str\" om, K. J}{1965}{Optimal Control of Markov Processes
 with Incomplete State Information.}{Journal of Mathematical Analysis
 and Applications}{Vol. 10, pp. 174--205}


\dissertation{Atkeson, Andrew G}{1988}{Essays in Dynamic International
Economics}{Stanford University}


 \article{Atkeson, Andrew G}{1991}{International Lending with Moral Hazard and
 Risk of Repudiation.}{Econometrica}{Vol. 59(4), pp. 1069--1089}


\article{Atkeson, Andrew, V. V. Chari, and Patrick J. Kehoe}{2010}{Sophisticated Monetary Policies.}{Quarterly Journal of Economics}{Vol. 125, pp. 47--89}


 \article{Atkeson, Andrew, and Robert E. Lucas, Jr}{1992}{On Efficient
 Distribution with Private Information.}{Review of Economic Studies}
 {Vol. 59(3), pp. 427--453}


 \article{Atkeson, Andrew, and Robert E. Lucas, Jr}{1995}{Efficiency
 and Equality in a Simple Model of Efficient Unemployment Insurance.}
 {Journal of Economic Theory}{Vol. 66(1), pp. 64--88}


 \contribution{Atkeson, Andrew, and Christopher Phelan}{1994}
 {Reconsidering the Costs of Business Cycles with Incomplete Markets}
 {Stanley Fischer and Julio J. Rotemberg}{NBER Macroeconomics Annual}
 {Cambridge, Mass.: MIT Press, pp. 187--207}




\contribution{Attanasio, Orazio P}{2000}{Consumption}
{John Taylor and Michael Woodford}{Handbook of Macroeconomics}
{Amsterdam: North-Holland}



 \article{Attanasio, Orazio P.,
 and Steven J. Davis}{1996}{Relative Wage Movements
 and the Distribution of Consumption.}{Journal of Political Economy}{Vol.
 104(6), pp. 1227--1262}


 \article{Attanasio, Orazio P., and Guglielmo Weber}{1993}{Consumption Growth,
 the Interest Rate and Aggregation.}{Review of Economic Studies}
 {Vol. 60(3), pp. 631--649}


 \book{Auerbach, Alan J., and Laurence J. Kotlikoff}{1987}{Dynamic Fiscal
 Policy}{New York: Cambridge University Press}


 \article{Auernheimer, Leonardo}{1974}{The Honest Government's Guide to the
 Revenue from the Creation of Money.}{Journal of Political Economy}
 {Vol. 82, pp. 598--606}


 \book{Azariadis, Costas}{1993}{Intertemporal Macroeconomics}
 {Cambridge, Mass.: Blackwell Press}

\article{Backus, David K,  Patrick J. Kehoe,  and , Finn E. Kydland}{1992}
{International Real Business Cycles.}{Journal of Political Economy}{Vol. 100(4), pp. 745-75}

\mimeo{Backus, David K. and Stanley E. Zin}{1994}{Reverse
Engineering the Yield Curve.}{NBER  Working Paper No. 4676}


 \article{Balasko, Y., and Karl Shell}{1980}{The Overlapping-Generations
 Model I: The Case of Pure Exchange without Money.}{Journal of
 Economic Theory}{Vol. 23(3), pp. 281--306}

 \mimeo{Barillas, F., A. Bhandari,  S. Bigio,   R.  Colacito,  M. Juillard, S. Kitao, C. Matthes, T. J. Sargent, and Y. Shin}{2012}
 {Practicing Dynare}{New York University}

 \article{Barillas, Francisco, Lars Peter  Hansen, and Thomas J. Sargent}{2009}{Doubts or variability?}
 {Journal of Economic Theory}{Vol. 144(6), pp. 2388-2418}



 \article{Barro, Robert J}{1974}{Are Government Bonds Net Wealth?}{Journal of
 Political Economy}{Vol. 82(6), pp. 1095--1117}


\article{Barro, Robert J}{1979}{On the Determination of Public Debt.}
{Journal of Political Economy}{Vol. 87, pp. 940--971}


 \article{Barro, Robert J., and David B. Gordon}{1983a}{A Positive Theory
 of Monetary Policy in a Natural Rate Model.}{Journal of Political Economy}
 {Vol. 91, pp. 589--610}


 \article{Barro, Robert J., and David B. Gordon}{1983b}{Rules, Discretion,
 and Reputation in a Model of Monetary Policy.}{Journal of Monetary
 Economics}{Vol. 12, pp. 101--121}


\book{Barro, Robert J., and Xavier Sala-i-Martin}{1995}{Economic Growth}
{New York: McGraw-Hill}


\article{Barseghyan, Levon,  Francesca Molinari, Ted O'Donoghue, and   Joshua C. Teitelbaum}{2013}
{The Nature of Risk Preferences: Evidence from Insurance Choices.}
% {Working Paper 11-03, Cornell University, Center for Analytic Economics}
% Levon Barseghyan & Francesca Molinari & Ted O'Donoghue & Joshua C. Teitelbaum, 2013.
%"The Nature of Risk Preferences: Evidence from Insurance Choices,"
{American Economic Review}{Vol. 103(6), pp. 2499-2529}


 \article{Barsky, Robert B., Gregory N. Mankiw, and Stephen P. Zeldes}{1986}
 {Ricardian Consumers with Keynesian Propensities.}{American Economic Review}
 {Vol. 76(4), pp. 676--691}


 \book{Basar, Tamer, and Geert Jan Olsder}{1982}{Dynamic Noncooperative Game
 Theory}{New York: Academic Press}




\article{Bassetto, Marco}{2002}{A Game-Theoretic View of the
Fiscal Theory of the Price Level.}{Econometrica}{Vol. 70(6),
pp. 2167--2195}


\article{Bassetto, Marco}{2005}{Equilibrium and Government
Commitment.}{Journal of Economic Theory}{Vol. 124, No. 1, pp. 79--105}



 \article{Bassetto, Marco, and Narayana Kocherlakota}{2004}{On the Irrelevance of Government Debt When Taxes Are Distortionary.}
 {Journal of Monetary Economics}{Vol. 51(2), pp. 299--304}

 \article{Baumol, William J}{1952}{The Transactions Demand for Cash:
 An Inventory Theoretic Approach.}{Quarterly Journal of Economics}
 {Vol. 66, pp. 545--556}

\article{Beker, Pablo, and Subir Chattopadhyay}{2010}
 {Consumption Dynamics in General Equilibrium: A Characterization When Markets Are Incomplete.}
 {Journal of Economic Theory}{Vol. 145, pp. 2133--2185}

 \book{Bellman, Richard}{1957}{Dynamic Programming}
{Princeton, N.J.: Princeton University Press}


 \book{Bellman, Richard, and Stuart E. Dreyfus}{1962}{Applied Dynamic Programming}
 {Princeton, N.J.: Princeton University Press}


\article{Benassy, Jean-Pascal}{1998}{Is There Always Too Little Research in
Endogenous Growth with Expanding Product Variety?}
{European Economic Review}{Vol. 42, pp. 61--69}

\book{Benassy, Jean-Pascal}{2011}{Macroeconomic Theory}{Oxford: Oxford University Press}

 \article{Benoit, Jean-Pierre, and Vijay Krishna}{1985}{Finitely Repeated
 Games.}{Econometrica}{Vol. 53, pp. 905--922}

 \article{Ben-Porath, Yoram}{1967}{The Production of Human Capital and
the Life Cycle of Earnings.}{Journal of Political Economy}
{Vol. 75, pp. 352--365}


 \article{Benveniste, Lawrence, and Jose Scheinkman}{1979}{On the
 Differentiability of the Value Function in Dynamic Models
 of Economics.}{Econometrica}{Vol. 47(3), pp. 727--732}


 \article{Benveniste, Lawrence, and Jose Scheinkman}{1982}{Duality Theory
 for Dynamic Optimization Models of Economics: The Continuous Time Case.}
 {Journal of Economic Theory}{Vol. 27, pp. 1--19}


 \article{Bernheim, B. Douglas, and Kyle Bagwell}{1988}{Is Everything Neutral?}
 {Journal of Political Economy}{Vol. 96(2), pp. 308--338}


\article{Bertola, Giuseppe, and Ricardo J. Caballero}{1994}{Cross-Sectional
Efficiency and Labour Hoarding in a Matching Model of Unemployment.}
{Review of Economic Studies}{Vol. 61, pp. 435--456}


 \book{Bertsekas, Dimitri P}{1976}{Dynamic Programming and Stochastic Control}
 {New York: Academic Press}


 \book{Bertsekas, Dimitri P}{1987}{Dynamic Programming: Deterministic
 and Stochastic Models}{Englewood Cliffs, N.J.: Prentice-Hall}


 \book{Bertsekas, Dimitri P., and Steven E. Shreve}{1978}{Stochastic Optimal
 Control: The Discrete Time Case}{New York: Academic Press}


\article{Bewley, Truman F}{1977}{The Permanent Income Hypothesis:
A Theoretical Formulation.}{Journal of Economic Theory}
{Vol. 16(2), pp. 252--292}


 \contribution{Bewley, Truman F}{1980}{The Optimum Quantity of Money}
 {J. H. Kareken and N. Wallace}{Models of Monetary Economies}
 {Minneapolis: Federal Reserve Bank of Minneapolis, pp. 169--210}


 \article{Bewley, Truman F}{1983}{A Difficulty with the Optimum
 Quantity of Money.}{Econometrica}{Vol. 51, pp. 1485--1504}


 \contribution{Bewley, Truman F}{1986}{Stationary Monetary Equilibrium with a
 Continuum of Independently Fluctuating Consumers}{Werner Hildenbrand and
 Andreu Mas-Colell}{Contributions to Mathematical Economics in Honor of
 Gerard Debreu}{Amsterdam: North-Holland, pp. 79--102}


 \article{Bhandari, Anmol, David G. Evans, Mikhail Golsov, and Thomas J. Sargent}{2017}{Fiscal Policy and Debt Management with Incomplete Markets.}
 {Quarterly Journal of Economics}{XXXXX}



\article{Black, Fisher, and Myron Scholes}{1973}{The Pricing of Options and
Corporate Liabilities.}{Journal of Political Economy}{Vol. 81, pp. 637--654}


 \article{Blackwell, David}{1965}{Discounted Dynamic Programming.}{Annals of
 Mathematical Statistics}{Vol. 36(1), pp. 226--235}


\article{Blanchard, Olivier Jean}{1985}{Debt, Deficits, and Finite
Horizons.}{Journal of Political Economy}
{Vol. 93(2), pp. 223--247}




\book{Blanchard, Olivier Jean, and Stanley Fischer}{1989}{Lectures on
Macroeconomics}{Cambridge, Mass.: MIT Press}




 \article{Blanchard, Olivier Jean, and Charles M. Kahn}{1980}{The Solution
 of Linear Difference Models under Rational Expectations.}
 {Econometrica}{Vol. 48(5), pp. 1305--1311}


\article{Blume, Lawrence  and  David Easley}{2006}
{If You're so Smart, why Aren't You Rich? Belief Selection in Complete and Incomplete Markets.}
 {Econometrica}{Vol. 74(4), pp. 929-966}


\article{Blundell, Richard and Ian Preston}{1998}{Consumption
Inequality and Income Uncertainty.}{Quarterly Journal of
Economics} {Vol. 113(2), pp. 603-640}






 \article{Bohn, Henning}{1995}{The Sustainability of Budget Deficits in a
 Stochastic Economy.}{Journal of Money, Credit, and Banking}{Vol. 27(1),
 pp. 257--271}


 \article{Bond, Eric W., and Jee-Hyeong Park}{2002}{Gradualism in Trade Agreements
 with Asymmetric Countries.}{Review of Economic Studies}{Vol. 69,
 pp. 379--406}


\article{Braun, R. Anton}{1994}
 {Tax Disturbances and Real Economic Activity in the Postwar United States.}
{Journal of Monetary Economics}{Vol. 33(3), pp. 441--462}




 \article{Breeden, Douglas T}{1979}{An Intertemporal Asset Pricing Model
 with Stochastic Consumption and Investment Opportunities.}
 {Journal of Financial Economics}{Vol. 7(3), pp. 265--296}


 \article{Brock, William A}{1972}{On Models of Expectations Generated by
 Maximizing Behavior of Economic Agents Over Time.}{Journal of
 Economic Theory}{Vol. 5, pp. 479--513}


 \article{Brock, William A}{1974}{Money and Growth: The Case of
 Long Run Perfect Foresight.}{International Economic Review}
 {Vol. 15, pp. 750--777}


 \ocontribution{Brock, William A}{1982}{Asset Prices in a Production Economy}
 {J. J. McCall}{The Economics of Information and Uncertainty.}
 {Chicago: University of Chicago Press, pp. 1--43}


 \contribution{Brock, William A}{1990}{Overlapping Generations Models
 with Money and Transactions Costs}{B. M. Friedman and F. H. Hahn}
 {Handbook of Monetary Economics, Vol. 1}{Amsterdam: North-Holland,
 pp. 263-295}


 \article{Brock, William A., and Leonard Mirman}{1972}{Optimal Economic
Growth and Uncertainty: The Discounted Case.}{Journal of Economic Theory}
{Vol. 4(3), pp. 479--513}




\contribution{Browning, Martin, Lars P. Hansen, and James J. Heckman}
{1999}{Micro Data and General Equilibrium Models}
{John Taylor and Michael Woodford}{Handbook of Macroeconomics. {\rm Volume 1A}}
{Amsterdam: North-Holland, pp. 543--633}




 \article{Bruno, Michael, and Stanley Fischer}{1990}{Seigniorage, Operating
 Rules, and the High Inflation Trap.}{Quarterly Journal of Economics}
 {Vol. 105, pp. 353--374}


 \article{Bryant, John, and Neil Wallace}{1984}{A Price Discrimination
 Analysis of Monetary Policy.}{Review of Economic Studies}
 {Vol. 51(2), pp. 279--288}

\article{Buera, Francisco, and  Juan Pablo Nicolini}{2004}
{Optimal maturity of government debt without state contingent bonds}{Journal of Monetary Economics}{Vol. 51(3), pp. 531-554}



 \article{Buiter, Willem H}{2002}{The Fiscal Theory of the Price Level:
A Critique.}{Economic Journal}{Vol. 112, pp. 459--480}


\article{Bulow, Jeremy, and Kenneth Rogoff}{1989}{Sovereign Debt:
 Is to Forgive to Forget?}{American Economic Review}
 {Vol. 79, pp. 43--50}


\article{Burdett, Kenneth, and  Kenneth L. Judd}{1983}
{Equilibrium Price Dispersion}{Econometrica}{Vol. 51(4), pp.  955-69}

\article{Burdett, Kenneth, and Dale T.  Mortensen}{1998}{Wage Differentials, Employer Size, and Unemployment}
 {International Economic Review}{Vol. 39(2), pp. 257--73}




\article{Burdett, Kenneth, Shouyong Shi, and Randall Wright}{2001}
{Pricing and Matching with Frictions.}{Journal of Political Economy}
{Vol. 109(5), pp. 1060--1085}







\article{Burnside, C., M. Eichenbaum, and S. Rebelo}{1993}
{Labor Hoarding and the Business Cycle.}{Journal of Political Economy}
{Vol. 101(2), pp. 245--273}


\article{Burnside, C., and M. Eichenbaum}{1996a}{Factor Hoarding and the
 Propagation of Business Cycle Shocks.}{American Economic Review}
{Vol. 86(5), pp. 1154--74}


\article{Burnside, C., and M. Eichenbaum}{1996b}{Small Sample Properties
of GMM Based Wald Tests.}{Journal of Business and Economic Statistics}
{Vol. 14(3), pp. 294--308}




\article{Caballero, Ricardo J}{1990}{Consumption Puzzles and Precautionary
Saving.}{Journal of Monetary Economics}{Vol. 25, No. 1, pp. 113-136}




 \ocontribution{Cagan, Phillip}{1956}{The Monetary Dynamics of Hyperinflation}
 {Milton Friedman}{Studies in the Quantity Theory of Money.}
 {Chicago: University of Chicago Press, pp. 25--117}


 \article{Calvo, Guillermo A}{1978}{On the Time Consistency of Optimal Policy
 in a Monetary Economy.}{Econometrica}{Vol. 46(6), pp. 1411--1428}

\article{Calvo, Guillermo A}{1983}{Staggered prices in a utility-maximizing framework.}{Journal of Monetary Economics}
 {Vol. 12, No.3, pp. 383--398}



\book{Campbell, John Y., Andrew W. Lo, and A. Craig MacKinlay}{1997}
{The Econometrics of Financial Markets}{Princeton, N.J.: Princeton University
Press}


 \article{Campbell, John Y., and John H. Cochrane}{1999}{By Force of Habit:
 A Consumption-Based Explanation of Aggregate Stock Market Behavior.}
 {Journal of Political Economy}{Vol. 107(2), pp. 205--251}


 \article{Campbell, John Y., and Robert Shiller}{1988}{The
 Dividend-Price Ratio And Expectations of Future Dividends and
 Discount Factors.}{Review of Financial Studies}{Vol. 1, pp.
 195--227}


 \article{Campbell, John Y., and Robert Shiller}{1991}
{Yield Spreads and Interest Rate Movements: A Bird's Eye View.}
{Review of Economic Studies}{Vol. 58, pp. 495--514}


\book{Canova, Fabio}{2007}{Method for Applied Macroeconomic Research}
{Princeton, New Jersey: Princeton University Press}

 \article{Carroll, Christopher D., and Miles S. Kimball}{1996}
 {On the Concavity of the Consumption Function.}{Econometrica}
 {Vol. 64(4), pp. 981--992}


 \article{Casella, Alessandra, and Jonathan S. Feinstein}{1990}
 {Economic Exchange during Hyperinflation.}{Journal of Political Economy}
 {Vol. 98(1), pp. 1--27}


 \article{Cass, David}{1965}{Optimum Growth in an Aggregative Model
 of Capital Accumulation.}{Review of Economic Studies}
 {Vol. 32(3), pp. 233--240}





\article{Cass, David}{1972}{On Capital Overaccumulation in the
Aggregative, Neoclassical Model of Economic Growth: A Complete
Characterization.}{Journal of  Economic Theory}{Vol.  4, pp.
200--233}

\article{Cass, David, and M. E. Yaari}{1966}{A Re-examination of the
 Pure Consumption Loans Model.}{Journal of Political Economy}
 {Vol. 74, pp. 353--367}

\article{Cass, David, and Karl Shell}{1983}{Do Sunspots Matter?}
{Journal of Political Economy}{Vol. 91(2), pp. 193--227}



 \article{Chamberlain, Gary, and Charles Wilson}{2000}{Optimal Intertemporal
 Consumption Under Uncertainty.}{Review of Economic Dynamics}
{Vol. 3(3), pp. 365--395}


 \article{Chamley, Christophe}{1986}{Optimal Taxation of Capital Income
 in General Equilibrium with Infinite Lives.}{Econometrica}
 {Vol. 54(3), pp. 607--622}


 \article{Chamley, Christophe, and Heraklis Polemarchakis}{1984}
 {Assets, General Equilibrium, and the Neutrality of Money.}
 {Review of Economic Studies}{Vol. 51, pp. 129--138}


\book{Champ, Bruce, and Scott Freeman}{1994}
{Modeling Monetary Economies}{New York: Wiley}


\article{Chang, Roberto}{1998}{Credible Monetary Policy in an Infinite
Horizon Model: Recursive Approaches.}{Journal of Economic Theory}
{Vol. 81(2), pp. 431--461}


\article{Chang, Yongsung, and Sun-Bin Kim}{2007}{Heterogeneity and
Aggregation: Implications for Labor-Market Fluctuations.}
{American Economic Review}{Vol. 97, pp. 1939--1956}

\article{Chari, V. V., Lawrence J. Christiano, and Martin Eichenbaum}
{1998}{Expectations Traps.}{Journal of Economic Theory}
{Vol. 81(2), pp. 462--492}




\article{Chari, V. V., Lawrence J. Christiano, and Patrick J. Kehoe}
 {1994}{Optimal Fiscal Policy in a Business Cycle Model.}
 {Journal of Political Economy}{Vol. 102(4), pp. 617--652}


 \article{Chari, V. V., Lawrence J. Christiano, and Patrick J. Kehoe}{1996}
 {Optimality of the Friedman Rule in Economies with Distorting Taxes.}
 {Journal of Monetary Economics}{Vol. 37(2), pp. 203--223}


 \article{Chari, V. V., and Patrick J. Kehoe}{1990}{Sustainable Plans.}
 {Journal of Political Economy}{Vol. 98, pp. 783--802}


 \article{Chari, V. V., and Patrick J. Kehoe}{1993a}{Sustainable Plans
 and Mutual Default.}{Review of Economic Studies}
 {Vol. 60, pp. 175--195}


 \article{Chari, V. V., and Patrick J. Kehoe}{1993b}{Sustainable Plans and
 Debt.}{Journal of Economic Theory}{Vol. 61, pp. 230--261}


 \ocontribution{Chari, V. V., Patrick J. Kehoe, and Edward C. Prescott}{1989}
 {Time Consistency and Policy}{Robert Barro}{Modern Business Cycle
 Theory.}{Cambridge, Mass.: Harvard University Press, pp. 265--305}


\article{Chatterjee, Satyajit, and Dean Corbae}{1996}{Money and Finance
with Costly Commitment.}{Journal of Monetary Economics}
{Vol. 37(2), pp. 225--248}



\article{Chen, Ren-Raw, and Louis Scott}{1993}{Maximum Likelihood
Estimation for a Multifactor Equilibrium Model of the Term Structure
of Interest Rates.}{Journal of Fixed Income}{pp. 95--09}



\article{Ch\'eron, Arnaud, Jean-Olivier Hairault, and Fran\c cois Langot}
{2013}{Life-Cycle Equilibrium Unemployment.}{Journal of Labor Economics}
{Vol. 31(4), pp. 843--882}




 \article{Cho, In-Koo, and Akihiko Matsui}{1995}{Induction and the Ramsey
 Policy.}{Journal of Economic Dynamics and Control}
{Vol. 19(5--7), pp. 1113--1140}


 \book{Chow, Gregory}{1981}{Econometric Analysis by Control Methods}
{New York: Wiley}


 \book{Chow, Gregory}{1997}{Dynamic Economics: Optimization
 by the Lagrange Method}{New York: Oxford University Press}


 \article{Christiano, Lawrence J}{1990}{Linear-Quadratic Approximation and
 Value-Function Iteration: A Comparison.}{Journal of Business
 and Economic Statistics}{Vol. 8(1), pp. 99--113}


\article{Christiano, Lawrence J., and M. Eichenbaum}{1992}{Current Real
Business Cycle Theories and Aggregate Labor Market Fluctuations.}
{American Economic Review}{Vol. 82(3)}


\article{Christiano, Lawrence J., M. Eichenbaum, and C.
Evans}{2003}{Nominal Rigidities and the Dynamic Effects of a Shock
to Monetary Policy.}{Journal of Political Economy}{forthcoming}


\article{Christiano, Lawrence J., Martin S. Eichenbaum,
and Mathias Trabandt}{2016}{Unemployment and Business Cycles.}
{Econometrica}{Vol. 84(4), pp. 1523--1569}


\article{Christiano, Lawrence J., R. Motto, and M.
Rostagno}{2003}{The Great Depression and the Friedman-Schwartz
Hypothesis.}{Proceedings, Federal Reserve Bank of Cleveland}{pp. 1119-1215}


\book{Christensen, Bent Jasper and Nicholas M. Kiefer}{2009}
{Economic Modeling and Inference}{Princeton, New Jersey: Princeton University
Press}



 \article{Clower, Robert}{1967}{A Reconsideration of the Microfoundations
 of Monetary Theory.}{Western Economic Journal}{Vol. 6, pp. 1--9}


 \article{Cochrane, John H}{1991}{A Simple Test of Consumption Insurance.}
 {Journal of Political Economy}{Vol. 99(5), pp. 957--976}




 \article{Cochrane, John H}{1997}{Where Is the Market Going? Uncertain Facts and
 Novel Theories.}{Economic Perspectives}{Vol. 21(6), pp. 3--37}


\contribution{Cochrane, John H., and Lars Peter Hansen}{1992}
{Asset Pricing Explorations for Macroeconomics}{Olivier Jean Blanchard and Stanley Fischer}
{NBER Macroeconomics Annual}{Cambridge, Mass.: MIT Press, pp. 115--165}

\article{Cochrane, John H}{2005}{Money as stock}{Journal of Monetary Economics}{Vol. 52(3), pp. 501--528}



 \article{Cogley, Timothy}{1999}{Idiosyncratic Risk and the Equity Premium: Evidence from the Consumer Expenditure Survey.}{Journal of Monetary Economics}{Vol. 49(2), pp.  309--334}

\article{Cogley, Timothy, Thomas J. Sargent, Viktor Tsyrennikov}{2013}{Wealth Dynamics in a Bond Economy with
Heterogeneous Beliefs}{Economic Journal}{XXXXX In press}







 \mimeo{Cole, Harold L., and Narayana Kocherlakota}{1998}
 {Dynamic Games with Hidden Actions and Hidden States.}{Federal Reserve Bank
 of Minneapolis: Staff Report  254}

\article{Cole, Harold L., and Narayana Kocherlakota}{2001}
 {Efficient Allocations with Hidden Income and Hidden Storage.}
 {Review of Economic-Studies}{Vol. 68(3), pp. 523--42}


\mimeo{Correia, Isabel, Emmanuel Farhi, Juan Pablo Nicolini, and Pedro Teles}{2010}
{Policy at the Zero Bound.}{Federal Reserve Bank of Minneapolis}

%\mimeo{Acemoglu, Daron}{1997}{Good Jobs versus Bad Jobs: Theory and Some
%Evidence.}{CEPR Discussion Paper No. 1588}

 \article{Constantinides, George M., and Darrell Duffie}{1996}{Asset Pricing
 with Heterogeneous Consumers.}{Journal of Political Economy}
 {Vol. 104(2), pp. 219--240}


 \book{Cooley, Thomas F}{1995}{Frontiers of Business Cycle Research}
 {Princeton, N.J.: Princeton University Press}


 \book{Cooper, Russell W}{1999}{Coordination Games: Complementarities
 and Macroeconomics}{New York: Cambridge University Press}


 \article{Correia, Isabel H}{1996}{Should Capital Income Be Taxed
 in the Steady State?}{Journal of Public Economics}
 {Vol. 60(1), pp. 147--151}


 \article{Correia, Isabel, and Pedro Teles}{1996}{Is the Friedman Rule Optimal
 When Money Is an Intermediate Good?}{Journal of Monetary Economics}
 {Vol. 38, pp. 223--244}


 \article{Cox, John C., Jonathan E. Ingersoll, Jr., and Stephen A. Ross}{1985a}
 {An Intertemporal General Equilibrium Model of Asset Prices.}
 {Econometrica}{Vol. 53(2), pp. 363--384}


 \article{Cox, John C., Jonathan E. Ingersoll, Jr., and Stephen A. Ross}{1985b}
 {A Theory of the Term Structure of Interest Rates.}{Econometrica}
 {Vol. 53(2), pp. 385--408}






\article{Dai, Qiang, and Kenneth J. Singleton}
{2000}{Specification Analysis of Affine Term Structure Models.}
{Journal of Finance}{Vol. LV, pp. 1943-1978}




\contribution{Davies, James B. and Anthony Shorrocks}{2000}{The Distribution of Wealth}
{A.B. Atkinson, F. Bourguignon}
{Handbook of Income Distribution, Vol. 1.}{Amsterdam: North-Holland, pp. 605--675}


 \mimeo{Davis, Steven J}{1995}{The Quality Distribution of Jobs and the
 Structure of Wages in Search Equilibrium.}{Chicago: University of Chicago Graduate School of Business}


 \book{Deaton, Angus}{1992}{Understanding Consumption}
 {New York: Oxford University Press}

 \article{Deaton, Angus and  Christina Paxson}{1994}{Intertemporal Choice and Inequality.}
 {Journal of Political Economy}{Vol. 102(3), pp. 437-67}




\article{Debreu, Gerard}{1954}{Valuation Equilibrium and Pareto Optimum.}
{Proceedings of the National Academy of Sciences}{Vol. 40, pp. 588--592}


 \book{Debreu, Gerard}{1959}{Theory of Value}{New York: Wiley}

\book{DeJong, David and Chetan Dave}{2011}{Structural Macroeconometrics, second edition}
{Princeton and Oxford: Princeton University Press}




\article{Den Haan, Wouter J., and Albert Marcet}{1990}
{Solving the Stochastic Growth Model by Parametering Expectations.}
 {Journal of Business and Economic Statistics}{Vol. 8(1), pp. 31--34}




 \article{Den Haan, Wouter J and Albert Marcet}{1994}{Accuracy in
 Simulations.}
 {Review of Economic Studies}{Vol.
61(1), pp.  3--17}




\article{Diamond, Peter A}{1965}{National Debt in a Neoclassical Growth
Model.}{American Economic Review}{Vol. 55, pp. 1126--1150}


 \article{Diamond, Peter A}{1981}{Mobility Costs, Frictional Unemployment, and
 Efficiency.}{Journal of Political Economy}{Vol. 89(4), pp. 798--812}


 \article{Diamond, Peter A}{1982}{Wage Determination and Efficiency in
 Search Equilibrium.}{Review of Economic Studies}{Vol. 49, pp. 217--227}


 \article{Diamond, Peter A}{1984}{Money in Search Equilibrium.}
 {Econometrica}{Vol. 52, pp. 1--20}


 \article{Diamond, Peter A., and Joseph Stiglitz}{1974}{Increases in Risk
 and in Risk Aversion.}{Journal of Economic Theory}
 {Vol. 8(3), pp. 337--360}


 \article{Diaz-Gim\' enez, J., Edward C. Prescott, T. Fitzgerald, and
 Fernando Alvarez}{1992}{Banking in Computable General Equilibrium Economies.}
 {Journal of Economic Dynamics and Control}{Vol. 16, pp. 533--560}


\article{Diaz-Gim\' enez, J., V. Quadrini, and J.V. R\'\i
os-Rull}{1997}{Dimensions of Inequality: Facts on the U.S.
Distributions of Earnings, Income, and Wealth.}{Quarterly Review}
{Federal Reserve Bank of Minneapolis, Vol. 21, pp. 3--21}


 \article{Dixit, Avinash K., Gene Grossman, and Faruk Gul}{2000}
 {The Dynamics of Political Compromise.}{Journal of Political Economy}{Vol. 108(3), pp. 531--568}


\article{Dixit, Avinash K., and Joseph E. Stiglitz}{1977}{Monopolistic
Competition and Optimum Product Diversity.}{American Economic Review}
{Vol. 67, pp. 297--308}

\article{De Santis, Massimiliano}{2007}{Individual Consumption Risk and the Welfare Cost
of Business Cycles}{American Economic Review}{Vol. 97, No. 4, pp. 1488--1506}


\article{Dolmas, J}{1998}{Risk Preferences and the Welfare Cost of
Business Cycles.}{Review of Economic Dynamics}
  {Vol. 1(3), pp. 646--676}

\mimeo{Domeij, David, and Jonathan Heathcote}{2000}{Capital versus
Labor Income Taxation with Heterogeneous Agents.}{Stockholm School of Economics}


 \book{Doob, Joseph L}{1953}{Stochastic Processes}{New York: Wiley}


 \article{Dornbusch, Rudiger}{1976}{Expectations and Exchange Rate Dynamics.}
 {Journal of Political Economy}{Vol. 84, pp. 1161--1176}


 \article{Dow, James R., Jr., and Lars J. Olson}{1992}{Irreversibility
 and the Behavior of Aggregate Stochastic Growth Models.}
 {Journal of Economic Dynamics and Control}{Vol. 16, pp. 207--233}


\article {Duffie, D. and L. G. Epstein}{1992} {Stochastic
Differential Utility.} {Econometrica} {Vol. 60(2), pp. 353--394}

 \book{Duffie, Darrell}{1996}{Dynamic Asset Pricing Theory}
 {Princeton, N.J.: Princeton University Press,  pp. xvii, 395}




\article{Duffie, Darrell, and Rui Kan}{1996}{A Yield-Factor Model of
Interest Rates.}{Mathematical Finance}{Vol. 6(4), pp. 379--406}


\article{Duffie, Darrell, J. Geanakoplos, A. Mas-Colell, and A.
McLennan}{1994}{Stationary Markov Equilibria.}{Econometrica}
{Vol. 62(4), pp. 745--781}

\article{Duffie, Darrell and Wayne Shafer}{1985}{Equilibrium in incomplete markets, I : A basic model of generic existence.}
 {Journal of Mathematical Economics}{Vol. 14(3), pp. 285--300}



\article{Eichenbaum, Martin}{1991}{Real Business-Cycle Theory: Wisdom
or Whimsy?}{Journal of Economic Dynamics and Control}{Vol. 15(4),
pp. 607--626}


 \article{Eichenbaum, Martin, and Lars P. Hansen}{1990}{Estimating Models with
 Intertemporal Substitution Using Aggregate Time Series Data.}
 {Journal of Business and Economic Statistics}{Vol. 8, pp. 53--69}


 \mimeo{Eichenbaum, Martin, Lars P. Hansen, and S.F. Richard}{1984}
 {The Dynamic Equilibrium Pricing of Durable Consumption Goods.}
 {Pittsburgh: Carnegie-Mellon University}


 \book{Elliott, Robert J., Lakhdar Aggoun, and John B. Moore}
 {1995}{Hidden Markov Models: Estimation and Control}
 {New York: Springer-Verlag}


\article {Ellsberg, Daniel}{1961} {Risk, Ambiguity and the Savage
Axioms} {Quarterly Journal of Economics} {Vol. 75(4), pp. 643--669}


\article{Engle, Robert F., and C. W. J.
Granger}{1987}{Co-Integration and Error Correction:
Representation, Estimation, and Testing.}{Econometrica}{Vol. 55(2), pp. 251--276}


 \article{Epstein, Larry G., and Stanley E. Zin}{1989}{Substitution, Risk
 Aversion, and the Temporal Behavior of Consumption and Asset Returns:
 A Theoretical Framework.}{Econometrica}{Vol. 57(4), pp. 937--969}


 \article{Epstein, Larry G., and Stanley E. Zin}{1991}{Substitution, Risk
 Aversion, and the Temporal Behavior of Consumption and Asset Returns: An
 Empirical Analysis.}{Journal of Political Economy}{Vol. 99(2), pp. 263--286}


\article{Ethier, Wilfred J}{1982}{National and International Returns
to Scale in the Modern Theory of International Trade.}
{American Economic Review}{Vol. 72, pp. 389--405}


%
%\ocontribution{Friedman, Milton}{1969}{The Optimum Quantity of Money}
%{Milton Friedman}{The Optimum Quantity of Money and Other Essays.}
%{Chicago: Aldine, pp. 1--50}

\ocontribution{Evans, David and Thomas J. Sargent}{2013}{History Dependent Public Policies}
{Thomas J. Sargent and Jouko Vilmunen}{Macroeconomics at the Service of Public Policy}{Oxford: Oxford University Press, pp. 120--140}

\book{Evans, George W.,  and Seppo Honkapohja}{2001}
{Learning and Expectations in Macroeconomics}{Princeton University Press, Princeton, New Jersey}


\article{Evans, George W., and Seppo Honkapohja}{2003}{Expectations
and the Stability Problem for Optimal Monetary Policies.} {Review
of Economic Studies}{Vol. 70(4), pp. 807--24}


 \article{Faig, Miquel}{1988}{Characterization of the Optimal Tax
 on Money When It Functions as a Medium of Exchange.}
 {Journal of Monetary Economics}{Vol. 22(1), pp. 137--148}


 \book{Fama, Eugene F}{1976a}{Foundations of Finance: Portfolio Decisions
 and Securities Prices}{New York: Basic Books}


 \article{Fama, Eugene F}{1976b}{Inflation Uncertainty and Expected Returns
 on Treasury Bills.}{Journal of Political Economy}{Vol. 84(3), pp. 427--448}


 \book{Farmer, Roger E. A}{1993}{The Macroeconomics of Self-fulfilling
 Prophecies}{Cambridge, Mass.: MIT Press}

\article{Fernandes, Ana and Christopher Phelan}{2000}{A Recursive Formulation for Repeated Agency with History Dependence}
{Journal of Economic Theory}{Vol. 91(2), pp.  223-247}



 \article{Fischer, Stanley}{1983}{A Framework for Monetary and Banking
 Analysis.}{Economic Journal}{Vol. 93, Supplement, pp. 1--16}


 \book{Fisher, Irving}{1913}{The Purchasing Power of Money: Its
 Determination and Relation to Credit, Interest and Crises}
 {New York: Macmillan}


 \book{Fisher, Irving}{[1907] 1930}{The Theory of Interest}
 {London: Macmillan}


 \article{Fisher, Jonas}
 {2006}{The Dynamic Effects of Neutral and Investment-Specific Technology Shocks.}
  {Journal of Political Economy}{Vol. 114, No. 3, pp. 413--451}


 {2003}{Technology Shocks Matter.}{Federal
 Reserve Bank of Chicago, Working Paper No. 14}



\article{Foley, Duncan K. and Martin F. Hellwig}{1975}
{Asset Management with Trading Uncertainty.}{Review of Economic Studies}{Vol. 42, no. 3, pp. 327-346}



\article{Frankel, Marvin}{1962}{The Production Function in Allocation and
Growth: A Synthesis.}{American Economic Review}{Vol. 52, pp. 995--1022}


\book{Friedman, Milton}{1953}{Essays in Positive Economics}{Chicago, Illinois: University of Chicago Press}

 \book{Friedman, Milton}{1957}{A Theory of the Consumption Function}
 {Princeton, N.J.: Princeton University Press}





 \article{Friedman, Milton}{1967}{The Role of Monetary Policy.}{American
 Economic Review}{Vol. 58, 1968, pp. 1--15.  Presidential Address
 delivered at the 80th Annual Meeting of the American Economic
 Association, Washington, DC, December 29}


\ocontribution{Friedman, Milton}{1969}{The Optimum Quantity of Money}
{Milton Friedman}{The Optimum Quantity of Money and Other Essays.}
{Chicago: Aldine, pp. 1--50}


 \book{Friedman, Milton, and Anna J. Schwartz}{1963}{A Monetary History
 of the United States, 1867--1960}{Princeton, N.J.: Princeton University Press
 and NBER}


\article{Fuchs, William Martin, and Francesco  Lippi}{2006}{Monetary Union with Voluntary Participation}{Review of Economic Studies}
 {Vol. 73, pp. 437-457}

 \article{Fudenberg, Drew, Bengt Holmstr\"om, and Paul Milgrom}{1990}
 {Short-Term Contracts and Long-Term Agency Relationships.}
 {Journal of Economic Theory}{Vol. 51(1)}

\article{Fudenberg Drew, and David M. Kreps}{1993}{Learning Mixed Equilibria}
{Games and Economic Behavior}{Vol. 5(3), pp. 320--367}



\book{Fudenberg, Drew and David K. Levine}{1998}
{The Theory of Learning in Games}{MIT Press, Cambridge, MA.}




 \book{Gabel, R. A., and R. A. Roberts}{1973}{Signals and Linear Systems}
 {New York: Wiley}


\article{Gale, David}{1973}{Pure Exchange Equilibrium of Dynamic Economic
Models.}{Journal of Economic Theory}{Vol. 6, pp. 12-36}


\article{Gali, Jordi}{1991}{Budget Constraints and Time-Series Evidence
on Consumption.}{American Economic Review}{Vol. 81(5), pp. 1238--1253}

\book{Gali, Jordi}{2008}{Monetary Policy, Inflation, and the Business Cycle: An Introduction to the New Keynesian Framework}
{Princeton and Oxford: Princeton University Press}


 \article{Gallant, R., L. P. Hansen, and G. Tauchen}{1990}
 {Using Conditional Moments of Asset Payoffs to Infer the Volatility of
 Intertemporal Marginal Rates of Substitution.}{Journal of Econometrics}
 {Vol. 45, pp. 145--179}

\article{Genicot, Garance and Debraj Ray}{2003}{Group Formation in
Risk-Sharing Arrangements.}{Review of Economic-Studies}{Vol. 70(1), pp. 87--113}


\book{Gittins, J.C}{1989}{Multi-armed Bandit and Allocation Indices}
{New York: Wiley}




\article{Golosov M., N. Kocherlakota, and A.  Tsyvinski}{2003}
{Optimal Indirect and Capital Taxation.}
{Review of Economic Studies}{Vol. 70, no. 3, July, pp. 569--587}






\article{Gomes, Joao, Jeremy Greenwood, and Sergio Rebelo}{2001}
{Equilibrium Unemployment.}{Journal of Monetary Economics}
{Vol. 48(1), pp. 109--152}


\mimeo{Gong, Frank F., and Eli M. Remolona}{1997}{A Three Factor
Econometric Model of the U.S. Term Structure.}{Federal Reserve
Bank of New York, Staff Report 19}


 \mimeo{Gourinchas, Pierre-Olivier, and Jonathan A. Parker}{1999}
 {Consumption over the Life Cycle.}{NBER Working Paper No.7271}


 \article{Granger, C. W. J}{1966}{The Typical Spectral Shape of an Economic
 Variable.}{Econometrica}{Vol. 34(1), pp. 150--161}


 \article{Granger, C. W. J}{1969}{Investigating Causal Relations by Econometric
 Models and Cross-Spectral Methods.}{Econometrica}{Vol. 37(3), pp. 424--438}
 %Reprinted in R.E. Lucas and T.J. Sargent, (eds.), {\sl Rational
 %Expectations and Econometric Practice}, University of Minnesota, 1981}


 \contribution{Green, Edward J}{1987}{Lending and the Smoothing of Uninsurable
 Income}{Edward C. Prescott and Neil Wallace}{Contractual Arrangements for
 Intertemporal Trade, Minnesota Studies in Macroeconomics series, Vol.
 1}{Minneapolis: University of Minnesota Press, pp. 3--25}


 \article{Green, Edward J., and Robert H. Porter}{1984}{Non-Cooperative
 Collusion under Imperfect Price Information.}{Econometrica}
 {Vol. 52, pp. 975--993}


\article{Greenwood, Jeremy, Zvi Hercowitz, and   Gregory W. Huffman}{1988}
{Investment, Capacity Utilization, and the Real Business Cycle.}{American Economic Review}{Vol. 78, no. 3, pp. 402-17}




\article{Greenwood, J., Z. Hercowitz, and P. Krusell}{1997}{Long
Run Implications of Investment-Specific Technological
Change.}{American Economic Review}{Vol. 78,   pp. 342--362}




\article{Grossman, Gene M., and Elhanan Helpman}{1991}{Quality Ladders in the
Theory of Growth.}{Review of Economic Studies}{Vol. 58, pp. 43--61}


 \article{Grossman, Sanford J., and Robert J. Shiller}{1981}{The Determinants of
 the Variability of Stock Market Prices.}{American Economic Review}
 {Vol. 71(2), pp. 222--227}




\mimeo{Gul, Faruk, and Wolfgang Pesendorfer}{2000}{Self-Control and the
Theory of Consumption.}{Princeton, New Jersey: Princeton University}


\mimeo{Guvenen, Fatih, Burhanettin Kuruscu, and Serdar Ozkan}{2011}
{Taxation of Human Capital and Wage Inequality: A Cross-Country Analysis}
{University of Minnesota, University of Toronto and Federal Reserve Board}

 \article{Guidotti, Pablo E., and Carlos A. Vegh}{1993}{The Optimal
 Inflation Tax When Money Reduces Transactions Costs: A Reconsideration.}
 {Journal of Monetary Economics}{Vol. 31(3), pp. 189--205}


\article{Hagedorn, Marcus, and Iourii Manovskii}{2008}{The Cyclical
Behavior of Equilibrium Unemployment and Vacancies Revisited.}
{American Economic Review}{Vol. 98(4), pp. 1692--1706}



 \article{Hall, Robert E}{1971}{The Dynamic Effects of Fiscal Policy in an
 Economy with Foresight.}{Review of Economic Studies}{Vol. 38, pp. 229--244}


 \article{Hall, Robert E}{1978}{Stochastic Implications of the Life
 Cycle-Permanent Income Hypothesis: Theory and Evidence.}{Journal of
 Political Economy}{Vol. 86(6), pp. 971--988. (Reprinted in {\it
 Rational Expectations and Econometric Practice}, ed.
 Robert E. Lucas, Jr., and Thomas J. Sargent, Minneapolis: University of Minnesota Press, 1981, pp. 501--520.)}


\article{Hall, Robert E}{1997}{Macroeconomic Fluctuations and the
Allocation of Time.}{Journal of Labor Economics}{Vol. 15(1),
pp. 223-250}



\article{Hall, Robert E.}{2005}{Employment Fluctuations with
Equilibrium Wage Stickiness.}
{American Economic Review}{Vol. 95(1), pp. 50--65}


\article{Hall, Robert E., and Dale Jorgenson}{1967}{Tax Policy and
Investment Behavior.}{The American Economic Review}{Vol. 57(3),
pp. 391--414}


\article{Hall, Robert E., and Paul R. Milgrom}{2008}{The
Limited Influence of Unemployment on the Wage Bargain.}
{American Economic Review}{Vol. 98(4), pp. 1653--1674}

 \book{Hamilton, James D}{1994}{Time Series Analysis}
 {Princeton, N.J.: Princeton University Press}


 \article{Hamilton, James D.,
 and Marjorie A. Flavin}{1986}{On the Limitations of
 Government Borrowing: A Framework for Empirical Testing.}
 {American Economic Review}{Vol. 76(4), pp. 808--819}


 \article{Hansen, Gary D}{1985}{Indivisible Labor and the Business Cycle.}
 {Journal of Monetary Economics}{Vol. 16, pp. 309--327}


 \article{Hansen, Gary D., and Ay\c se \.Imrohoro\u glu}{1992}{The Role of
 Unemployment Insurance in an Economy with Liquidity Constraints and Moral
 Hazard.}{Journal of Political Economy}{Vol. 100 (1), pp. 118--142}


 \article{Hansen, Lars P}{1982a}{Consumption, Asset Markets, and
 Macroeconomic Fluctuations: A Comment.}{Carnegie-Rochester Conference
 Series on Public Policy}{Vol. 17, pp. 239--250}


\article{Hansen, Lars P}{1982b}{Large Sample Properties of
Generalized Method of Moments Estimators.}{Econometrica}
{Vol. 50, pp. 1029--1060}


 \ocontribution{Hansen, Lars P., Dennis Epple, and Will Roberds}
 {1985}{Linear-Quadratic Duopoly Models of Resource Depletion}
 {Thomas J. Sargent}{Energy, Foresight, and Strategy.}
 {Washington, DC: Resources for the Future, pp. 101--142}


\contribution{Hansen, Lars P. and Thomas J. Sargent}{1991} {Exact
Linear Rational Expectations Models}{Lars P. Hansen and Thomas J.
Sargent}{Rational Expectations Econometrics}{Boulder, Colo.: Westview
Press, pp. 45--76}


\book{Hansen, Lars P. and Thomas J. Sargent}{2013}{Recursive Models of Dynamic Linear Economies}
{Princeton University Press, Princeton, New Jersey}

\article {Hansen, L. P., T. J. Sargent, and N. E. Wang}{2002}{Robust Permanent Income and Pricing with
Filtering.}{Macroeconomic Dynamics}{Vol. 6(1), pp. 40--84}

\article{Hansen L. P. and T. J. Sargent}{2001} {Robust Control and
Model Uncertainty.} {American Economic Review} {Vol. 91(2), pp.
60--66}

\book{Hansen, Lars P. and Thomas J. Sargent}{2008}{Robustness}
{Princeton University Press, Princeton, New Jersey}

 \article{Hansen, Lars P., and Ravi Jagannathan}{1991}{Implications
 of Security Market Data for Models of Dynamic Economies.}
 {Journal of Political Economy}{Vol. 99, pp. 225--262}


 \article{Hansen, Lars P., and Ravi Jagannathan}{1997}{Assessing Specification
 Errors in Stochastic Discount Factor Models.}{Journal of Finance}
 {Vol. 52(2), pp. 557--590}


\contribution{Hansen, Lars P., William T. Roberds, and Thomas J. Sargent}{1991}
{Time Series Implications of Present Value Budget Balance and of Martingale Models
of Consumption and Taxes}{L. P. Hansen and T. J. Sargent}{Rational Expectations
and Econometric Practice}{Boulder, Colo.: Westview Press, pp. 121--161}


 \article{Hansen, Lars P., and Thomas J. Sargent}{1980}{Formulating and Estimating Dynamic Linear Rational Expectations Models.}{Journal of Economic Dynamics and Control}{Vol. 2(1), pp. 7--46}

 \contribution{Hansen, Lars P., and Thomas J. Sargent}{1981}{Linear Rational
 Expectations Models for Dynamically Interrelated Variables}
 {R. E. Lucas, Jr., and T. J. Sargent}{Rational Expectations and Econometric
 Practice}{Minneapolis: University of Minnesota Press, pp. 127--156}


 \article{Hansen, Lars P., and Thomas J. Sargent}{1982}{Instrumental Variables
 Procedures for Estimating Linear Rational Expectations Models.}
 {Journal of Monetary Economics}{Vol. 9(3), pp. 263--296}





\article{Hansen, Lars P., and Thomas J. Sargent}{1995}{Discounted Linear Exponential Quadratic Gaussian Control.}{IEEE Transactions on Automatic Control}{Vol. 40, pp. 968--971}


\article{Hansen, Lars P., Thomas J. Sargent, and Thomas D. Tallarini, Jr}{1999}
{Robust Permanent Income and Pricing.}{Review of Economic Studies}
{Vol. 66(4), pp. 873--907}


\book{Hansen, Lars P. and Thomas J. Sargent}{2008}
{Robustness}{Princeton University Press, Princeton, New Jersey}

 \contribution{Hansen, Lars P. and Thomas J. Sargent}{2011}
 {Wanting Robustness in Macroeconomics}{Benjamin Friedman and Michael Woodford}{Handbook of Monetary Economics, Volume 3B}
{Elsevier, North-Holland, pp.
1097-1157}

\article{Hansen, Lars P. and Thomas J. Sargent}{2012}{Three Types of Ambiguity.}
{Journal of Monetary Economics}{Volume 59, Issue 5,  pp. 422-445}

 \article{Hansen, Lars P., and Kenneth J. Singleton}{1982}{Generalized
 Instrumental Variables Estimation of Nonlinear Rational Expectations
 Models.}{Econometrica}{Vol. 50(5), pp. 1269--1286}


 \article{Hansen, Lars P., and Kenneth J. Singleton}{1983}{Stochastic Consumption, Risk Aversion, and the Temporal Behavior of Asset Returns.}
 {Journal of Political Economy}{Vol. 91(2), pp. 249--265}






\book{Hansen, Lars P., and Thomas J. Sargent}{2013}{Recursive Models
of Dynamic Linear Economies}{Princeton, New Jersey: Princeton University Press}




\article{Harrison, J.  Michael  and David M. Kreps}{1978}
{Speculative Investor Behavior in a Stock Market with Heterogeneous Expectations}{Quarterly Journal of Economics}{Vol. 92(2), pp. 323--36}



\article{Harrison, J. Michael, and David M.  Kreps}{1979}{Martingales and Arbitrage
in Multiperiod Security Markets.}{Journal of Economic Theory}
{Vol. 20, pp. 381--408}

\article{Heathcote, Jonathan, Kjetil  Storesletten, and Giovanni L. Violante}{2008}{Insurance and opportunities: A welfare analysis of labor market risk.}
{Journal of Monetary Economics}{Vol. 55(3), pp. 501--525}

\article{Heathcote, Jonathan,  Fabrizio Perri, and  Giovanni L. Violante}{2010}
{Unequal We Stand: An Empirical Analysis of Economic Inequality in the United States: 1967-2006.}{Review of Economic Dynamics}
 {Vol. 13, No. 1, pp. 15--51}

\mimeo{Heathcote, Jonathan, Kjetil  Storesletten, and Giovanni L. Violante}{2012}{Consumption and Labor Supply with Partial
Insurance: An Analytical Framework.}{Working paper, Federal Reserve Bank of Minneapolis and NYU}



 \article{Heaton, John, and Deborah J. Lucas}{1996}{Evaluating the Effects
 of Incomplete Markets on Risk Sharing and Asset Pricing.}
 {Journal of Political Economy}{Vol. 104(3), pp. 443--487}


 \article{Helpman, Elhanan}{1981}{An Exploration in the Theory of Exchange-Rate
 Regimes.}{Journal of Political Economy}{Vol. 89(5), pp. 865--890}


 \article{Hirshleifer, Jack}{1966}{Investment Decision under Uncertainty:
 Applications of the State Preference Approach.}{Quarterly Journal of
 Economics}{Vol. 80(2), pp. 252--277}


\article{Holmstr\"om, Bengt}{1983}{Equilibrium Long Term Labour Contracts.}
{Quarterly Journal of Economics, Supplement}{Vol. 98(1), pp. 23--54}


\article{Hopenhayn, Hugo A., and Juan Pablo Nicolini}{1997}
 {Optimal Unemployment Insurance.}{Journal of Political Economy}
 {Vol. 105(2), pp. 412--438}




 \article{Hopenhayn, Hugo A., and Edward C. Prescott}{1992}{Stochastic
 Monotonicity and Stationary Distributions for Dynamic Economies.}
 {Econometrica}{Vol. 60(6), pp. 1387--1406}


 \article{Hopenhayn, Hugo, and Richard Rogerson}{1993}{Job Turnover and
 Policy Evaluation: A General Equilibrium Analysis.}{Journal of
 Political Economy}{Vol. 101, pp. 915--938}




 \article{Hosios, Arthur, J}{1990}{On the Efficiency of Matching and
 Related Models of Search and Unemployment.}{Review of Economic Studies}
 {Vol. 57, pp. 279--298}


 \article{Hubbard, R. Glenn, Jonathan Skinner, and Stephen P. Zeldes}
 {1995}{Precautionary Saving and Social Insurance.}{Journal of Political
 Economy}{Vol. 103(2), pp. 360--399}


 \article{Huffman, Gregory}{1986}{The Representative Agent, Overlapping
 Generations, and Asset Pricing.}{Canadian Journal of Economics}
 {Vol. 19(3), pp. 511--521}


 \article{Huggett, Mark}{1993}{The Risk Free Rate in Heterogeneous-Agent,
 Incomplete-Insurance Economies.}{Journal of Economic Dynamics
 and Control}{Vol. 17(5-6), pp. 953--969}




\mimeo{Huggett, Mark, and Sandra Ospina}{2000}{Aggregate Precautionary
Savings: When Is the Third Derivative Irrelevant?}{Washington, D.C.: Georgetown University}




 \article{\.Imrohoro\u glu, Ay\c se}{1992}{The Welfare Cost of Inflation
 Under Imperfect Insurance.}{Journal of Economic Dynamics and Control}
 {Vol. 16(1), pp. 79--92}




 \article{\.Imrohoro\u glu, Ay\c se, Selahattin \.Imrohoro\u glu, and
 Douglas Joines}{1995}{A Life Cycle Analysis of Social Security.}{Economic
 Theory}{Vol. 6(1), pp. 83--114}




\article{Ingram, Beth Fisher, Narayana Kocherlakota, and N.
E. Savin}{1994}{Explaining Business Cycles: A Multiple-Shock
Approach.}{Journal of Monetary Economics}{Vol. 34(3), pp,
415--428}




\article{Ireland, Peter N}{1994}{Inflationary Policy and Welfare
with Limited Credit Markets.}{Journal of Financial Intermediation}
{Vol. 3(3), pp. 245--271}


 \article{Ireland, Peter N}{1997}{Sustainable Monetary Policies.}
 {Journal of Economic Dynamics and Control}{Vol. 22, pp. 87--108}






\article{Jacobson, David H}{1973}{Optimal Stochastic Linear Systems
with Exponential Performance Criteria and Their Relation to Deterministic Differential Games.}{IEEE Transactions on Automatic Control}
{Vol. 18(2), pp. 124--131}


 \book{Johnson, Norman, and Samuel Kotz}{1971}{Continuous Univariate
 Distributions}{New York: Wiley}


\article{Jones, Charles I}{1995}{R\&D-Based Models of Economic Growth.}
{Journal of Political Economy}{Vol. 103, pp. 759--784}

\article{Jones, Larry E.}{2008}{A Note on the Joint Occurrence of
Insurance and Gambling}{Macroeconomic Dynamics}{Vol. 12, pp. 97-111}


\article{Jones, Larry E., and Rodolfo Manuelli}{1990}{A Convex Model
of Equilibrium Growth: Theory and Policy Implications.}
{Journal of Political Economy}{Vol. 98, pp. 1008--1038}


 \article{Jones, Larry E., and Rodolfo E. Manuelli}{1992}{Finite Lifetimes
 and Growth.}{Journal of Economic Theory}{Vol. 58, pp. 171--197}


\article{Jones, Larry E., Rodolfo E. Manuelli, and Peter E. Rossi}{1993}
{Optimal Taxation in Models of Endogenous Growth.}{Journal of
Political Economy}{Vol. 101, pp. 485--517}


 \article{Jones, Larry E., Rodolfo E. Manuelli, and Peter E. Rossi}{1997}
 {On the Optimal Taxation of Capital Income.}{Journal of Economic Theory}
 {Vol. 73(1), pp. 93--117}


 \article{Jovanovic, Boyan}{1979a}{Job Matching and the Theory of Turnover.}
 {Journal of Political Economy}{Vol. 87(5), pp. 972--990}


\article{Jovanovic, Boyan}{1979b}{Firm-Specific Capital and Turnover.}
{Journal of Political Economy}{Vol. 87(6), pp. 1246--1260}


\article{Jovanovic, Boyan, and Yaw Nyarko}{1996}{Learning by Doing
and the Choice of Technology.}{Econometrica}{Vol. 64(6), pp. 1299--1310}


\article{Judd, Kenneth L}{1985a}{On the Performance of Patents.}
{Econometrica}{Vol. 53, pp. 567--585}


\article{Judd, Kenneth L}{1985b}{Redistributive Taxation in a Simple Perfect
Foresight Model.}{Journal of Public Economics}{Vol. 28, pp. 59--83}


\mimeo{Judd, Kenneth L}{1990}{Cournot versus Bertrand: A Dynamic Resolution.}
{Stanford, Calif.: Hoover Institution, Stanford University.}


 \contribution{Judd, Kenneth L}{1996}{Approximation, Perturbation, and
Projection Methods in Economic Analysis}{Hans Amman, David Kendrick, and John
 Rust}{Handbook of Computational Economics, Vol. 1}{Amsterdam: North-Holland}


 \book{Judd, Kenneth L}{1998}{Numerical Methods in Economics}
 {Cambridge, Mass.: MIT Press}


\article{Judd, Kenneth L., Sevin Yeltekin, and  James Conklin}{2003}{Computing Supergame Equilibria.}
   {Econometrica}{Vol.  71, no. 4, pp. 1239--1254}

\mimeo{Judd, Kenneth L., and  Andrew Solnick}{1994}{Numerical dynamic
programming with shape preserving splines.}{Stanford, Calif.: Hoover Institution}




\mimeo{Kahn, Charles, and William Roberds}{1998}
{Real-Time Gross Settlement and the Costs of Immediacy.}
{Federal Reserve Bank of Atlanta, Working Paper No.98--21, December}


 \article{Kalman, R. E}{1960}{Contributions to the Theory of Optimal Control.}
 {Bol. Soc. Mat. Mexicana}{Vol. 5, pp. 102--119}


 \article{Kalman, R. E., and R. S. Bucy}{1961}{New Results in Linear Filtering
 and Prediction Theory.}{J. Basic Eng., Trans. ASME, Ser. D}
 {Vol. 83, pp. 95--108}


 \article{Kandori, Michihiro}{1992}{Repeated Games Played by
 Overlapping Generations of Players.}{Review of Economic Studies}
 {Vol. 59(1), pp. 81--92}


 \article{Kareken, John, T. Muench, and N. Wallace}{1973}
 {Optimal Open Market Strategy: The Use of Information Variables.}
 {American Economic Review}{Vol. 63(1), pp. 156--172}


\book{Kareken, John, and Neil Wallace}{1980}{Models of Monetary Economies}
 {Minneapolis: Federal Reserve Bank of Minneapolis, pp. 169--210}


 \article{Kareken, John, and Neil Wallace}{1981}{On the Indeterminacy of
 Equilibrium Exchange Rates.}{Quarterly Journal of Economics}{Vol. 96,
 pp. 207--222}


\article{Kaplan, Greg and  Giovanni L. Violante}{2010}
{How Much Consumption Insurance beyond Self-Insurance}{American Economic Journal: Macroeconomics}
{Vol. 2(4), pp. 53--87}
%{Vol. 2(4), pp. 53--87}


 \article{Kehoe, Patrick, and Fabrizio Perri}{2002}{International
 Business Cycles with Endogenous Incomplete Markets.}
 {Econometrica}{Vol. 70(3), pp. 907--928}


 \article{Kehoe, Timothy J., and David K. Levine}{1984}
 {Intertemporal Separability in Overlapping-Generations Models.}
 {Journal of Economic Theory}{Vol. 34, pp. 216--226}


 \article{Kehoe, Timothy J., and David K. Levine}{1985}{Comparative Statics
 and Perfect Foresight in Infinite Horizon Economies.}{Econometrica}
 {Vol. 53, pp. 433--453}


\article{Kehoe, Timothy J., and David K. Levine}{1993}{Debt-Constrained
Asset Markets.}{Review of Economic Studies}{Vol. 60(4), pp. 865--888}




\book{Keynes, John Maynard}{1940}{How to Pay for the War: A Radical
Plan for the Chancellor of the Exchequer}{London: Macmillan}




 \article{Kihlstrom, Richard E., and Leonard J. Mirman}{1974}
 {Risk Aversion with Many Commodities.}{Journal of Economic Theory}
 {Vol. 8, pp. 361--388}


 \book{Kim, Chang-Jin, and Charles R. Nelson}{1999}{State Space Models
 with Regime Switching}{Cambridge, Mass.: MIT Press}

\article{Kim, J., and S. Kim}{2003}{Spurious Welfare Reversals in
International Business Cycle Models.}{Journal of International
Economics}{Vol. 60, pp. 471--500}

\article{Kimball, Miles S., and Mankiw, Gregory}{1989}{Precautionary
Saving and the Timing of Taxes.}{Journal of Poltical Economy}
{Vol. 97(4), pp. 863--879}




 \article{Kimball, M. S}{1990}{Precautionary Saving in the Small and
 in the Large.}{Econometrica}{Vol. 58, pp. 53--73}


 \article{Kimball, M. S}{1993}{Standard Risk Aversion.}{Econometrica}
 {Vol. 63(3), pp. 589--611}


 \article{Kimbrough, Kent P}{1986}{The Optimum Quantity of Money Rule
 in the Theory of Public Finance.}{Journal of Monetary Economics}
 {Vol. 18, pp. 277--284}


 \article{King, Robert G., and Charles I. Plosser}{1988}{Real Business Cycles:
 Introduction.}{Journal of Monetary Economics}{Vol. 21, pp. 191--193}


 \article{King, Robert G., Charles I. Plosser, and Sergio T. Rebelo}{1988}
 {Production, Growth and Business Cycles: I. The Basic Neoclassical
 Model.}{Journal of Monetary Economics}{Vol. 21, pp. 195--232}


\ocontribution{King, Robert G., and Alexander L. Wolman}{1999}{What Should
the Monetary Authority Do When Prices are Sticky}{John B. Taylor}
{Monetary Policy Rules.}{Chicago: University of Chicago Press, pp. 349-398}


 \article{Kiyotaki, Nobuhiro, and Randall Wright}{1989}{On Money as a Medium of
 Exchange.}{Journal of Political Economy}{Vol. 97(4), pp. 927--954}


 \mimeo{Kiyotaki, Nobuhiro, and Randall Wright}{1990}{Search for a Theory of
 Money.}{Cambridge, Mass.: National Bureau of Economic Research, Working Paper No. 3482}


 \article{Kiyotaki, Nobuhiro, and Randall Wright}{1993}{A Search-Theoretic
 Approach to Monetary Economics.}{American Economic Review}
 {Vol. 83(1), pp. 63--77}


\article{Kletzer, Kenneth M. and Brian D. Wright}{2000}{Sovereign
Debt as Intertemporal Barter.}{American Economic Review}{Vol. 90(3), pp. 621--39}

 \article{Kocherlakota, Narayana R}{1996a}{The Equity Premium:
 It's Still a Puzzle.}{Journal of Economic Literature}
 {Vol. 34(1), pp. 42--71}


 \article{Kocherlakota, Narayana R}{1996b}{Implications of Efficient Risk Sharing without Commitment.}
 {Review of Economic Studies}{Vol. 63(4), pp. 595--609}


 \article{Kocherlakota, Narayana R}{1998}{Money Is Memory.}{Journal of
 Economic Theory}{Vol. 81 (2), pp. 232--251}


 \article{Kocherlakota, Narayana R., and Christopher Phelan}{1999}
 {Explaining the Fiscal Theory of the Price Level.}
 {Quarterly Review}{Federal Reserve Bank of Minneapolis, Vol. 23(4), pp. 14--23}



 \article{Kocherlakota, Narayana R., and Neil Wallace}{1998}
 {Incomplete Record-Keeping and Optimal Payment Arrangements.}
 {Journal of Economic Theory}{Vol. 81(2), pp. 272--289}




 \article{Kocherlakota, Narayana R}{2004}{Figuring Out the Impact of Hidden Savings on Optimal Unemployment Insurance.}{Review of Economic
 Dynamics.}{forthcoming}




 \mimeo{Koeppl, Thorsten V}{2003}{Differentiability of the Efficient Frontier
 when Commitment to Risk Sharing Is Limited}
 {Frankfurt, Germany: European Central Bank}






 \book{Koopmans, Tjalling C}{1965}{On the Concept of Optimal Growth}{The
 Econometric Approach to Development Planning}{Chicago: Rand McNally}


\article {Kreps, D. M. and E. L. Porteus}{1978}{Temporal Resolution
of Uncertainty and Dynamic Choice.}{Econometrica}{Vol. 46(1), pp.
185--200}

\mimeo{Kreps, David M}{1979}{Three Essays on Capital Markets.}{Technical
Report 298. Stanford, Calif.: Institute for Mathematical Studies in the Social Sciences, Stanford University}


 \book{Kreps, David M}{1988}{Notes on the Theory of Choice}{Boulder, Colo.: Westview Press}


 \book{Kreps, David M}{1990}{Game Theory and Economic Analysis}{New York:
Oxford University Press}




\mimeo{Krueger, Dirk}{1999}{Risk Sharing in Economies with Incomplete
Markets.}{Stanford, California: Stanford University}


\article{Krueger, Dirk, and Fabrizio Perri}{2004}% 2003a
{On the Welfare
Consequences of the Increase in Inequality in the United States.}
{NBER Macroeconomics Annual 2003}{Volume 18, pp. 83--138}



\article{Krueger, Dirk, and Fabrizio Perri}{2006}% {2003b}
{Does Income Inequality Lead to Consumption Inequality? Evidence and Theory.}
{Review of Economic Studies}{Vol. 73(1), pp. 163--193}


\article{Krueger, Dirk,  Fabrizio Perri, Luigi Pistaferri, and Giovanni L. Violante}{2010}
{Cross Sectional Facts for Macroeconomists.}{Review of Economic Dynamics}{Vol. 13(1), pp. 1--14}


\article{Krusell, Per, Toshihiko Mukoyama, Richard Rogerson,
Ay\c sg\" ul \c Sahin}
{2008}{Aggregate Implications of Indivisible Labor, Incomplete Markets, and
Labor Market Frictions.}{Journal of Monetary Economics}{Vol. 55, pp. 961--979}

 \article{Krusell, Per, and Anthony Smith}{1998}{Income and Wealth
 Heterogeneity in the Macroeconomy.}{Journal of Political Economy}
 {Vol. 106(5), pp. 867--896}


 \book{Kwakernaak, Huibert, and Raphael Sivan}{1972}{Linear Optimal Control
 Systems}{New York: Wiley}


 \article{Kydland, Finn E., and Edward C. Prescott}{1977}{Rules Rather than
 Discretion: The Inconsistency of Optimal Plans.}{Journal of Political
 Economy}{Vol. 85(3), pp. 473--491}


 \article{Kydland, Finn E., and Edward C. Prescott}{1980}{Dynamic Optimal
 Taxation, Rational Expectations and Optimal Control.}{Journal of
 Economic Dynamics and Control}{Vol. 2(1), pp. 79--91}


 \article{Kydland, Finn E., and Edward C. Prescott}{1982}{Time to Build and
 Aggregate Fluctuations.}{Econometrica}{Vol. 50(6), pp. 1345--1371}






 \article{Labadie, Pamela}{1986}{Comparative Dynamics and Risk Premia in an
 Overlapping Generations Model.}{Review of Economic Studies}
 {Vol. 53(1), pp. 139--152}



\article{Lagos, Ricardo}{2000}{An Alternative Approach to
 Search Frictions.}{Journal of Political Economy}{108(5), pp. 851--873}


\mimeo{Laibson, David I}{1994}{Hyperbolic Discounting and Consumption.}
{Cambridge, Mass.: Massachusetts Institute of Technology}


\article{Davig, Troy,  Eric M. Leeper, and Todd B. Walker}{2010}{``Unfunded liabilities'' and uncertain fiscal financing.}
 {Journal of Monetary Economics}{Vol. 57, No. 5, pp.  600--619}



 \article{Leland, Hayne E}{1968}{Saving and Uncertainty: The Precautionary
 Demand for Saving.}{Quarterly Journal of Economics}{Vol. 82(3), pp.
465--473}


 \dissertation{LeRoy, Stephen F}{1971}{The Determination of Stock Prices}
 {Philadelphia: University of Pennsylvania}


 \article{LeRoy, Stephen F}{1973}{Risk Aversion and the Martingale Property of
 Stock Prices.}{International Economic Review}{Vol. 14(2), pp. 436--446}


 \article{LeRoy, Stephen F}{1982}{Risk Aversion and the Term Structure of
 Interest Rates.}{Economics Letters}{Vol. 10(3--4), pp. 355--361.
(Correction in {\it Economics Letters}, 1983, Vol. 12(3--4); 339--340.)}


 \article{LeRoy, Stephen F}{1984a}{Nominal Prices and Interest Rates in
 General Equilibrium: Money Shocks.}{Journal of Business}
 {Vol. 57(2), pp. 177--195}


 \article{LeRoy, Stephen F}{1984b}{Nominal Prices and Interest Rates in
 General Equilibrium: Endowment Shocks.}{Journal of Business}
 {Vol. 57(2), pp. 197--213}


 \article{LeRoy, Stephen F., and Richard D. Porter}{1981}{The Present-Value
 Relation: Tests Based on Implied Variance Bounds.}{Econometrica}
 {Vol. 49(3), pp. 555--574}


\article{Lettau, Martin and Sydney Ludvigson}{2001}{Consumption,
Aggregate Wealth, and Expected Stock Returns.}{Journal of Finance}
{Vol. LVI(3), pp. 815--849}


\article{Lettau, Martin and Sydney Ludvigson}{2004}{Understanding
Trend and Cycle in Asset Values: Reevaluating the Wealth Effect on
Consumption.}{American Economic Review}{Vol. 94(1), pp. 276--299}

 \article{Levhari, David, and Leonard J. Mirman}{1980}{The Great Fish War:
 An Example Using a Dynamic Cournot-Nash Solution.}
 {Bell Journal of Economics}{Vol. 11(1)}


 \article{Levhari, David, and T. N. Srinivasan}{1969}{Optimal Savings under
 Uncertainty.}{Review of Economic Studies}{Vol. 36(2), pp. 153--163}


\book{Levine, David K., and Drew Fudenberg}{1998}{The Theory of Learning
in Games}{Cambridge, Mass.: MIT Press}


\article{Levine, David K., and William R. Zame}{2002}
{Does Market Incompleteness Matter?}{Econometrica}{Vol. 70(5), pp.  1805-1839}





\article{Ligon, Ethan}{1998}{Risk Sharing and Information in Village Economies.}
{Review of Economic Studies}{Vol. 65(4), pp. 847--864}




\article{Ligon, Ethan, Jonathan P.\ Thomas, and Tim Worrall}{2002}
{Informal Insurance Arrangements with Limited Commitment: Theory and
Evidence from Village Economies.}
{Review of Economic Studies}{Vol. 69(1), pp. 209--244}




 \article{Lippman, Steven A., and John J. McCall}{1976}{The Economics of
 Job Search: A Survey.}{Economic Inquiry}{Vol. 14(3), pp. 347--368}


 \article{Ljungqvist, Lars}{2002}{How Do Layoff Costs Affect
Employment?}{Economic Journal}{Vol. 112, pp. 829--853}


\article{Ljungqvist, Lars, and Thomas J. Sargent}{1998}
{The European Unemployment Dilemma.}{Journal of Political Economy}
{Vol. 106, pp. 514--550}


 \contribution{Ljungqvist, Lars, and Thomas J. Sargent}{2005}{Lotteries
for Consumers versus Lotteries for Firms}{Timothy Kehoe, T.N. Srinivasan
and John Whalley}{Frontiers in Applied General Equilibrium Modeling: In
Honor of Herbert Scarf}{Cambridge University Press, pp 119--126}



\contribution{Ljungqvist, Lars, and Thomas J. Sargent}{2006}{Do Taxes
Explain European Employment?: Indivisible Labor, Human Capital, Lotteries,
and Savings}{Daron Acemoglu, Kenneth Rogoff, and Michael Woodford}
{Macroeconomics Annual 2006} {Cambridge MA.,  MIT Press pp 181--224}



\article{Ljungqvist, Lars, and Thomas J. Sargent}{2007}{Understanding
European Unemployment with Matching and Search-Island Models.}
{Journal of Monetary Economics}{Vol. 54, pp. 2139--2179}

\article{Ljungqvist, Lars, and Thomas J. Sargent}{2011}
{A Labor Supply Elasticity Accord?}{American Economic Review}
{Vol. 101, pp. 487--491}


\article{Ljungqvist, Lars, and Thomas J. Sargent}{2014}
{Career Length: Effects of Curvature of Earnings Profiles,
Earnings Shocks, Taxes, and Social Security.}
{Review of Economic Dynamics}{Vol. 17, pp. 1--20}


\article{Ljungqvist, Lars, and Thomas J. Sargent}{2017}{The
Fundamental Surplus.}{American Economic Review}{Vol. 107, pp. XXXXX }




 \article{Lucas, Robert E., Jr}{1972}{Expectations and the Neutrality
 of Money.}{Journal of Economic Theory}{Vol. 4, pp. 103--124}


 \article{Lucas, Robert E., Jr}{1973}{Some International Evidence on
 Output-Inflation Trade-Offs.}{American Economic Review}
 {Vol. 63, pp. 326--334}


 \contribution{Lucas, Robert E., Jr}{1976}{Econometric Policy Evaluation:
 A Critique}{K. Brunner and A. H. Meltzer}{The Phillips Curve and
 Labor Markets}{Amsterdam: North-Holland, pp. 19--46}


 \article{Lucas, Robert E., Jr}{1978}{Asset Prices in an Exchange Economy.}
 {Econometrica}{Vol. 46(6), pp. 1429--1445}


 \pcontribution{Lucas, Robert E., Jr}{1980a}{Equilibrium in a Pure Currency
 Economy}{J. H. Kareken and N. Wallace}{Economic Inquiry}{Vol. 18(2),
 pp. 203-220. (Reprinted in {\it Models of Monetary Economies},
 Federal Reserve Bank of Minneapolis, 1980, pp. 131--145.)}


 \article{Lucas, Robert E., Jr}{1980b}{Two Illustrations
of the Quantity Theory of Money.}
{American Economic Review}{Vol. 70, pp. 1005--1014}


 \ocontribution{Lucas, Robert E., Jr}{1981}{Econometric Testing of the Natural
 Rate Hypothesis}{Robert E. Lucas, Jr.}{Studies of Business-Cycle
 Theory.}{Cambridge, Mass.: MIT Press, pp. 90--103.  (Reprinted from {\sl The
 Econometrics of Price Determination Conference}, ed. Otto Eckstein.
 Washington, D.C.: Board of Governors of the Federal Reserve System, 1972,
 pp. 50--59)}


 \article{Lucas, Robert E., Jr}{1982}{Interest Rates and Currency Prices in a
 Two-Country World.}{Journal of Monetary Economics}{Vol. 10(3), pp.
 335--360}


\article{Lucas, Robert E. Jr.}{1986}{Principles of Fiscal and
Monetary Policy.}{Journal of Monetary Economics}{Vol. 17(1), pp.
117--134}

 \book{Lucas, Robert E., Jr}{1987}{Models of Business Cycles}{Yrjo Jahnsson
 Lectures Series. London: Blackwell}


\article{Lucas, Robert E., Jr}{1988}{On the Mechanics of Economic Development.}
{Journal of Monetary Economics}{Vol. 22, pp. 3--42}




\article{Lucas, Robert E., Jr.}{1990}{Supply-Side Economics: An
Analytical Review.}{Oxford Economic Papers}{Vol. 42(2), pp.
293-316}






\article{Lucas, Robert E., Jr}{1992}{On Efficiency and Distribution.}
{Economic Journal}{Vol. 102(4), pp. 233--247}



\article{Lucas, Robert E., Jr.}{2003}{Macroeconomic
Priorities.}{American Economic Review}{93(1), pp 1--14}


 \article{Lucas, Robert E., Jr., and Edward C. Prescott}{1971}{Investment under
 Uncertainty.}{Econometrica}{Vol. 39(5), pp. 659--681}


\article{Lucas, Robert E., Jr., and Edward C. Prescott}{1974}
{Equilibrium Search and Unemployment.}{Journal of Economic Theory}
{Vol. 7(2), pp. 188--209}


 \article{Lucas, Robert E., Jr., and Nancy Stokey}{1983}{Optimal Monetary and
 Fiscal Policy in an Economy without Capital.}{Journal of Monetary
 Economics}{Vol. 12(1), pp. 55--94}






\book{Luenberger, David G}{1969}{Optimization by Vector Space
Methods}{New York: Wiley}


\mimeo{Lustig, Hanno}{2000}{Secured Lending and Asset Prices.}
{Stanford, Calif.: Stanford University, Department of Economics}


\mimeo{Lustig, Hanno}{2003}{The Market Price of Aggregate Risk and
the Wealth Distribution}{Chicago: University of
Chicago, Department of Economics}


\article{Mace, Barbara}{1991}{Full Insurance in the Presence of
Aggregate Uncertainty.}{Journal of Political Economy}{Vol. 99(5), pp. 928--956}


 \article{Mankiw, Gregory N}{1986}{The Equity
Premium and the Concentration of Aggregate Shocks.}{Journal of
Financial Economics}{Vol. 17(1), pp. 211--219}




\article{Mankiw, N. Gregory,  and Stephen P. Zeldes}{1991}{The Consumption of Stockholders and
Nonstockholders.}{Journal
of Financial Economics}{Vol. 29(1), pp.  97--112}




 \article{Manuelli, Rodolfo, and Thomas J. Sargent}{1988}{Models of Business
 Cycles: A Review Essay.}{Journal of Monetary Economics}
 {Vol. 22 (3), pp. 523--542}


\article{Manuelli, Rodolfo and Thomas J. Sargent}{2010}{Alternative Monetary
Policies in a Turnpike Economy.}{Macroeconomic Dynamics}{Vol. 14, No. 5, pp. 727-762}


\mimeo{Manuelli, Rodolfo, Ananth Seshadri, and Yongseok Shin}{2012}
{Lifetime Labor Supply and Human Capital Investment.}{Washington
University in St.\ Louis and University of Wisconsin}

\article{Marcet, Albert,   Francesc Obiols-Homs, and Philippe Weil}{2007}
{Incomplete markets, labor supply and capital accumulation.}{ Journal of Monetary Economics}{Vol. 54, no. 8, pp. 2621--2635}



 \article{Marcet, Albert, and Ramon Marimon}{1992}{Communication, Commitment, and Growth.}{Journal of Economic Theory}{Vol. 58(2), pp. 219--249}


 \mimeo{Marcet, Albert, and Ramon Marimon}{1999}{Recursive Contracts.}
 {Universitat Pompeu Fabra, Barcelona}


\article{Marcet, Albert, and Juan Pablo Nicolini}{1999}{Recurrent
Hyperinflations and Learning.}{American Economic Review}{Vol. 93, No. 3, pp. 1476--1498}




\contribution{Marcet, Albert, and Thomas J. Sargent}{1989}
{Least Squares Learning and the Dynamics of Hyperinflation}
{William Barnett, John Geweke, and Karl Shell}{Economic Complexity:
Chaos, Sunspots, and Nonlinearity}{New York: Cambridge University Press}

\article{Marcet, Albert, and Thomas J. Sargent}{1989}{Convergence of least squares learning mechanisms in self-referential linear stochastic models.}
{Journal of Economic Theory}{Vol. 48, number 2, pp. 337-368} %August.




%\mimeo{Marcet, Albert, Thomas J. Sargent, and Juha Sepp\"al\"a}
%{1996}{Optimal Taxation without State-Contingent Debt.}
%{Universitat Pompeu Fabra and Stanford University}


\article{Marcet, Albert, and Kenneth J. Singleton}{1999}{Equilibrium
Asset Prices and Savings of Heterogeneous Agents in the Presence of
Incomplete Markets and Portfolio Constraints.}{Macroeconomic Dynamics}
{Vol. 3(2), pp. 243--277}


%\ocontribution{Marimon, Ramon}{forthcoming}
%{The Fiscal Theory of Money as an Unorthodox Financial
%Theory of the Firm}{Axel Leijonhufvud}{Monetary Theory
%as a Basis for Monetary Policy.}{International Economic Association (IEA)}


 \article{Marimon, Ramon, and Shyam Sunder}{1993}{Indeterminacy of Equilibria
 in a Hyperinflationary World: Experimental Evidence.}{Econometrica}
 {Vol. 61(5), pp. 1073--1107}


\article{Marimon, Ramon, and Fabrizio Zilibotti}{1999}{Unemployment vs.
Mismatch of Talents: Reconsidering Unemployment Benefits.}
{Economic Journal}{Vol. 109, pp. 266--291}

\book{Majumdar, Mukul}{2009}{Equilibrium, Welfare, and Uncertainty: Beyond Arrow-Debreu}
{London: Routledge}

 \book{Mas-Colell, Andrew, Michael D. Whinston, and Jerry R. Green}
 {1995}{Microeconomic Theory}{New York: Oxford University Press}


 \article{Matsuyama, Kiminori, Nobuhiro Kiyotaki, and Akihiko Matsui}{1993}
 {Toward a Theory of International Currency.}{Review of Economic Studies}
 {Vol. 60(2), pp. 283--307}


 \article{McCall, John J}{1970}{Economics of Information and Job Search.}
 {Quarterly Journal of Economics}{Vol. 84(1), pp. 113--126}




\article{McCall, B. P}{1991}{A Dynamic Model of Occupational Choice.}
{Journal of Economic Dynamics and Control}{Vol. 15(2), pp. 387--408}


 \article{McCallum, Bennett T}{1983}{The Role of Overlapping-Generations
 Models in Monetary Economics.}{Carnegie-Rochester Conference
 Series on Public Policy}{Vol. 18, pp. 9--44}


 \article{McCallum, Bennett T}{2001}{Indeterminacy, Bubbles, and the
Fiscal Theory of Price Level Determination.}{Journal of Monetary Economics}
{Vol. 47(1), pp. 19--30}



\book{McCandless, George T., and Neil Wallace}{1992}{Introduction to Dynamic
Macroeconomic Theory: An Overlapping Generations Approach}{Cambridge, Mass.:
Harvard University Press}


\article{McGrattan, Ellen R}{1994a}{A Note on Computing Competitive
Equilibria in Linear Models.}{Journal of Economic Dynamics and Control}
{Vol. 18(1), pp. 149--160}


\article{McGrattan, Ellen R}{1994b}{The Macroeconomic Effects of Distortionary Taxation.}
{Journal of Monetary Economics}{Vol. 33, pp. 573--601}


\article{McGrattan, Ellen R}{1996}{Solving the Stochastic Growth Model with
a Finite Element Method.}{Journal of Economic Dynamics and Control}
{Vol. 20(1--3), pp. 19--42}


 \article{Mehra, Rajnish, and Edward C. Prescott}{1985}{The Equity Premium: A
 Puzzle.}{Journal of Monetary Economics}{Vol. 15(2), pp. 145--162}


\article{Mendoza, Enrique G, and Linda L. Tesar}{1998}
{The International Ramifications of Tax Reforms: Supply-Side Economics in a Global Economy.}
{American Economic Review}{Vol. 88(1), pp. 226-45}



\article{Menzio, Guido, and Shouyong Shi}{2011}
{Efficient Search on the Job and the Business Cycle.}
{Journal of Political Economy}{Vol. 119(3), pp. 468--510}

\article{Menzio, Guido, Irina A. Telyukova, and Ludo Visschers}{2016}
{Directed Search over the Life Cycle.}
{Review of Economic Dynamics}{Vol. 19, pp. 38--62}


 \article{Mertens, Karel, and  Morten O. Ravn}{2011}
{Understanding the Aggregate Effects of Anticipated and Unanticipated Tax Shocks.}
{Review of Economic Dynamics}{Vol. 14(1), pp.  27--54}



\mimeo{Miao, Jianjun}{2003}{Competitive Equilibria of Economies with a Continuum of Consumers and Aggregate Shocks}
{Boston:  Boston University, Department of Economics}

 \article{Miller, Bruce L}{1974}{Optimal Consumption with a Stochastic
 Income Stream.}{Econometrica}{Vol. 42(2), pp. 253--266}


\article{Miller, Marcus and Mark Salmon}{1985}{Dynamic Games and
the Time Inconsistency of Optimal Policy in Open Economies.}
{Economic Journal}{Vol. 95, pp. 124-37}

\article{Miller, Robert A}{1984}{Job Matching and Occupational
Choice.}{Journal of Political Economy}{Vol. 92(6), pp. 1086--1120}


\book{Miranda, Mario J., and Paul L. Fackler}{2002}{Applied
Computational Economics and Finance}{Cambridge Mass.: MIT Press}


 \ocontribution{Modigliani, Franco, and Richard Brumberg}{1954}
 {Utility Analysis and the Consumption Function: An Interpretation
 of Cross-Section Data}{K. K. Kurihara}{Post-Keynesian Economics.}
 {New Brunswick, N.J.: Rutgers University Press}




 %Monetary and Fiscal Policy, edited by Torsten Persson and Guido Tabellini.
 %Vol. 1: Credibility; Vol. 2:  Politics.  MIT Press, Cambridge,
 %Massachussetts, (1994).




\article{Modigliani, F., and M. H. Miller}{1958}{The Cost of Capital, Corporation Finance, and the Theory of Investment.}{American Economic Review}{Vol. 48(3), pp. 261--297}


 \article{Moen, Espen R}{1997}{Competitive Search Equilibrium.}
 {Journal of Political Economy}{Vol. 105(2), pp. 385--411}


\article{Montgomery, James D}{1991}{Equilibrium Wage Dispersion and
Involuntary Unemployment.}{Quarterly Journal of Economics}
{Vol. 106, pp. 163--179}


 \ocontribution{Mortensen, Dale T}{1982}{The
Matching Process as a Noncooperative
 Bargaining Game}{John J. McCall}{The Economics of Information and
 Uncertainty.}{Chicago: University of Chicago Press,
 for the National Bureau of Economic Research, pp. 233--258}


 \article{Mortensen, Dale T}{1994}{The Cyclical Behavior of Job
 and Worker Flows.}{Journal of Economic Dynamics and Control}
 {Vol. 18, pp. 1121--1142}


 \article{Mortensen, Dale T., and Christopher A. Pissarides}{1994}
 {Job Creation and Job Destruction in the Theory of Unemployment.}
 {Review of Economic Studies}{Vol. 61, pp. 397--415}


\contribution{Mortensen, Dale T., and Christopher A. Pissarides}{1999a}
{New Developments in Models of Search in the Labor Market}
{Orley Ashenfelter and David Card}{Handbook of Labor Economics,
Vol. 3B}{Amsterdam: Elsevier/North-Holland}


\article{Mortensen, Dale T., and Christopher A. Pissarides}{1999b}{Unemployment
Responses to `Skill-Biased' Technology Shocks: The Role of Labour
Market Policy.}{Economic Journal}{Vol. 109, pp. 242--265}

\article{Mulligan, Casey B}{2001}{Aggregate Implications of Indivisible
Labor}{Advances in Macro\-economics}{Vol. 1(1): Article 4}

 \contribution{Muth, J. F}{[1960] 1981}{Estimation of Economic Relationships
 Containing Latent Expectations Variables}{R. E. Lucas, Jr.,
 and T. J. Sargent}{Rational Expectations and Econometric Practice}
 {Minneapolis: University of Minnesota Press, pp. 321--328}


 \article{Muth, John F}{1960}{Optimal Properties of Exponentially Weighted
 Forecasts.}{Journal of the American Statistical Association}
 {Vol. 55, pp. 299-306}


 \article{Muth, John F}{1961}{Rational Expectations and the Theory of Price
 Movements.}{Econometrica}{Vol. 29, pp. 315--335}






 \book{Naylor, Arch, and George Sell}{1982}{Linear Operator Theory in
 Engineering and Science}{New York: Springer}


\article{Neal, Derek}{1999}{The Complexity of Job Mobility among
Young Men.}{Journal of Labor Economics}{Vol. 17(2), pp. 237--261}




\article{Negishi, T.}{1960}{Welfare Economics and Existence of an
Equilibrium for a Competitive Economy.}{Metroeconomica}{Vol. 12, pp. 92--97}




 \contribution{Nerlove, Marc}{1967}{Distributed Lags and Unobserved Components
 in Economic Time Series}{William Fellner et al.}{Ten Economic Studies
 in the Tradition of Irving Fisher}{New York: Wiley}


 \article{Niepelt, Dirk }{2004}{The Fiscal Myth of the Price Level}{Quarterly Journal of Economics}{Vol. 119(1), pp. 276--299}



\book{Noble, Ben, and James W. Daniel}{1977}{Applied Linear Algebra}
{Englewood Cliffs, N.J.: Prenctice Hall}






 \article{O'Connell, Stephen A., and Stephen P. Zeldes}{1988}{Rational Ponzi
 Games.}{International Economic Review}{Vol. 29(3), pp. 431--450}

\article{Obstfeld, M}{1994} {Evaluating Risky Consumption Paths:
the Role of Intertemporal Substitutability.}
  {European Economic Review}
  {Vol. 38(7), pp. 1471--1486}

\article{Otrok, Christopher}{2001}{On Measuring the Welfare Cost
of Business Cycles.}{Journal of Monetary Economics}{Vol. 47(1), pp.
61--92}




\article{Paal, Beatrix}{2000}{Destabilizing Effects of a Successful Stabilization:
A Forward-Looking Explanation of the Second Hungarian Hyperinflation.}
{Journal of Economic Theory}{Vol. 15(3), pp. 599--630}


\article{Pavan, Ronni}{2011}
{Career Choice and Wage Growth.}
{Journal of Labor Economics}
{Vol. 29(3), pp. 549-587}


\article{Pavoni, Nicola and Giovanni  L. Violante}{2007}{Optimal Welfare-to-Work Programs.}
 {Review of Economic Studies}{Vol. 74, No. 1, pp. 283--318}


\article{Pearlman, J. G.}{1992}{Reputational and Nonreputational
Policies Under Partial Information.}{Journal of Economic Dynamics
and Control} {Vol. 16(2), pp. 339--358}

\article{Pearlman, J. G., D. A. Currie, and P. L. Levine}{1986}
{Rational expectations with partial information.} {Economic
Modeling}{Vol. 3, pp. 90-1-5}

 \article{Peled, Dan}{1984}{Stationary Pareto Optimality of Stochastic Asset
 Equilibria with Overlapping Generations.}{Journal of Economic Theory}
 {Vol. 34, pp. 396--403}

%
%\article{Persson, Mats, Torsten Persson, and Lars E. O. Svensson}{1988}
%{Time Consistency of Fiscal and Monetary Policy.}{Econometrica}
%{Vol. 55, pp. 1419--1432}


\article{Peters, Michael}{1991}{Ex Ante Price Offers in Matching
Games Non-Steady States.}{Econometrica}
{Vol. 59(5), pp. 1425--1454}

\article{Petrongolo, Barbara, and Christopher A. Pissarides}{2001}
{Looking into the Black Box: A Survey of the Matching Function.}
{Journal of Economic Literature}{Vol. 39(2), pp. 390--431}


 \article{Phelan, Christopher}{1994}{Incentives and Aggregate Shocks.}
 {Review of Economic Studies}{Vol. 61(4), pp. 681--700}


 \article{Phelan, Christopher, and Robert M. Townsend}{1991}
 {Computing Multi-period, Information-Constrained Optima.}
 {Review of Economic Studies}{Vol. 58(5), pp. 853--881}


\article{Phelan, Christopher, and Ennio Stacchetti}{2001}{Sequential
Equilibria in a Ramsey Tax Model.}{Econometrica}{Vol. 69(6), pp.
1491--1518}




%Sequential Equilibria in a Ramsey Tax Model AU:
%Phelan,-Christopher; Stacchetti,-Ennio AF: Federal Reserve Bank of
%Minneapolis; U MI SO: Econometrica-. November 2001; 69(6):
%1491-1518




 \book{Phelps, Edmund S}{1970}{Introduction to Microeconomic Foundations
 of Employment and Inflation Theory}{New York: Norton}


\article{Phelps, Edmund S.,  and Robert A. Pollak}{1968}{On Second-Best
National Saving and Game-Equilibrium Growth.}{Review of Economic
Studies}{Vol. 35(2), pp. 185--199}

\article{Piazzesi, Monika}{2005}{Bond Yields and the Federal Reserve}{Journal of Political Economy}{Vol. 113(2), pp. 311--344}

\contribution{Piazzesi, Monika, and Martin Schneider}{2006}{Equilibrium Yield Curves}{Daron Acemoglu, Kenneth Rogoff, and Michael Woodford}
{Macroeconomics Annual 2006} {Cambridge MA.,  MIT Press pp 389-442}

\mimeo{Piazzesi, Monika, Juliana Salomao, and Martin Schneider}{2015}
{Trend and Cycle in Bond Premia}{Stanford University}




%
% \contribution{Anderson, Evan W., Lars P. Hansen, Ellen R. McGrattan,
% and Thomas J. Sargent}{1996}{Mechanics of Forming and Estimating
% Dynamic Linear Economies}{Hans M. Amman, David A. Kendrick,
% and John Rust}{Handbook of Computational Economics, Vol. 1,
% Handbooks in Economics, Vol. 13}{Amsterdam:
% Elsevier Science, North-Holland, pp. 171--252}






\article{Pissarides, Christopher A}{1983}{Efficiency Aspects of
the Financing of Unemployment Insurance and Other Government Expenditures.}{Review of Economic Studies}{Vol. 50(1), pp. 57--69}


 \book{Pissarides, Christopher A}{1990}{Equilibrium Unemployment Theory}
 {Cambridge, U.K.: Basil Blackwell}


\article{Pissarides, Christopher A.}{1992}{Loss of Skill During
Unemployment and the Persistence of Employment Shocks.}
{Quarterly Journal of Economics}{Vol. 107(4), pp. 1371--1391}

\article{Pissarides, Christopher A.}{2009}{The Unemployment
Volatility Puzzle: Is Wage Stickiness the Answer?}
{Econometrica}{Vol. 77(5), pp. 1339--1369}

 \article{Pratt, John W}{1964}{Risk Aversion in the Small and in the Large.}
 {Econometrica}{Vol. 32(1--2), pp. 122--136}

\article{Prescott, Edward C}{1977}{Should Control Theory Be Used for Economic Stabilization?}
{1977}{Journal-of-Monetary-Economics. Supplementary Series}{Vol. 7, pp. 13--38}


\article{Prescott, Edward C}{2002} {Richard T. Ely Lecture:
Prosperity and Depression.} {American Economic Review} {Vol.
92(2), pp. 1--15}


\article{Prescott, Edward C}{2005a}
{The Transformation of Macroeconomic Policy and Research.}
{{\rm In} Les Prix Nobel 2004}{Stockholm: Almqvist \& Wiksell International, pp. 370--395.
http://nobelprize.org/economics/laureates/2004/prescott-lecture.pdf}


 \contribution{Prescott, Edward C.}{2005b}{Nonconvexities in Quantitative
General Equilibrium Studies of Business Cycles}{Timothy Kehoe, T.N. Srinivasan
and John Whalley}{Frontiers in Applied General Equilibrium Modeling:
In Honor of Herbert Scarf}{Cambridge University Press, pp 95--118}


\contribution{Prescott, Edward C}{2006a}{Comment}
{Daron Acemoglu, Kenneth Rogoff, and Michael Woodford}
{Macroeconomics Annual 2006} {Cambridge MA.,  MIT Press pp 233--242}



\article{Prescott, Edward C}{2006b}
{Nobel Lecture: The Transformation of Macroeconomic Policy and Research.}
{Journal of Political Economy}{Vol. 114 pp. 203--235}


 \article{Prescott, Edward C., and Rajnish Mehra}{1980}{Recursive Competitive
 Equilibrium: The Case of Homogeneous Households.}{Econometrica}
 {Vol. 48(6), pp. 1365--1379}


\article{Prescott, Edward C., Richard Rogerson, and Johanna Wallenius}{2009}
{Lifetime Aggregate Labor Supply with Endogenous Workweek Length.}
{Review of Economic Dynamics}{Vol. 12 pp. 23--36}


 \ocontribution{Prescott, Edward C., and Robert M. Townsend}{1980}
 {Equilibrium Under Uncertainty: Multiagent Statistical Decision
 Theory}{Arnold Zellner}{Bayesian Analysis in Econometrics and
 Statistics.}{Amsterdam: North-Holland, pp. 169--194}


 \article{Prescott, Edward C., and Robert M. Townsend}{1984a}
 {General Competitive Analysis in an Economy with Private Information.}
 {International Economic Review}{Vol. 25, pp. 1--20}


 \article{Prescott, Edward C., and Robert M. Townsend}{1984b}{Pareto Optima
 and Competitive Equilibria with Adverse Selection and Moral Hazard.}
 {Econometrica}{Vol. 52, pp. 21--45}


 \article{Putterman, Martin L., and Shelby Brumelle}{1979}{On the Convergence of
 Policy Iteration on Stationary Dynamic Programming.}{Mathematics of
 Operations Research}{Vol. 4(1), pp. 60--67}


 \article{Putterman, Martin L., and M. C. Shin}{1978}{Modified Policy Iteration
 Algorithms for Discounted Markov Decision Problems.}
 {Management Science}{Vol. 24(11), pp. 1127--1137}






\article{Quah, Danny}{1990}{Permanent and Transitory Movements in
Labor Income: An Explanation for `Excess Smoothness' in
Consumption.}{Journal of Political Economy}
{Vol. 98(3), pp. 449--475}






\article{Razin, Assaf, and Efraim Sadka}{1995}{The Status of Capital Income Taxation in the Open Economy.}
{FinanzArchiv}{Vol. 52(1), pp. 21--32}


\article{Rebelo, Sergio}{1991}{Long-Run Policy Analysis and Long-Run Growth.}
{Journal of Political Economy}{Vol. 99, pp. 500--521}


 \article{Reinganum, Jennifer F}{1979}{A Simple Equilibrium Model of Price
 Dispersion.}{Journal of Political Economy}{Vol. 87(4), pp. 851--858}


 \mimeo{R\'\i os-Rull, V\'\i ctor Jos\'e}{1994a}{Life-Cycle Economies
 and Aggregate Fluctuations.}{University of Pennsylvania}


 \mimeo{R\'\i os-Rull, V\'\i ctor Jos\'e}{1994b}{Population Changes and Capital
 Accumulation: The Aging of the Baby Boom.}{Philadelphia: University of Pennsylvania, Department of Economics}


\article{R\'\i os-Rull, V\'\i ctor Jos\'e}{1994c}{On the Quantitative Importance of Market Completeness.}{Journal of Monetary Economics}{Vol. 34(3), pp. 463--496}


 \ocontribution{R\'\i os-Rull, V\'\i ctor Jos\'e}{1995}{Models with Heterogeneous Agents}{Thomas F. Cooley}{Frontiers of Business Cycle Research.}{Princeton, N.J.: Princeton University Press, pp. 98--125}


\article{R\'\i os-Rull, V\'\i ctor Jos\'e}{1996}{Life-Cycle Economies and Aggregate Fluctuations.}{Review of Economic Studies}{Vol. 63(3), pp. 465--489}




\article{Rodriguez, Santiago Budria, Javier D\'iaz-Gim\'enez, Vincenzo Quadrini, and Jose-Victor R\'ios-Rull}
{2002}{Updated Facts on the U.S. Distributions of Earnings, Income, and Wealth.}
{Quarterly Review}{Federal Reserve Bank of Minneapolis, Summer}


\article{Roberds, William T}{1996}{Budget Constraints and Time-Series Evidence
on Consumption: Comment.}{American Economic Review}{Vol. 86(1), pp. 296--297}

\book{Robert, Christian P. and George Casella}{2004}{Monte Carlo Statistical Methods, Second Edition}
{Springer, New York}


\article{Rogerson, William P}{1985a}{Repeated Moral Hazard.}{Econometrica}
{Vol. 53(1), pp. 69--76}


\article{Rogerson, William P}{1985b}{The First Order Approach to
Principal-Agent Problems.}{Econometrica}
{Vol. 53(6), pp. 1357--1367}


 \article{Rogerson, Richard}{1988}{Indivisible Labor, Lotteries, and
 Equilibrium.}{Journal of Monetary Economics}{Vol. 21, pp. 3--16}
 
 \contribution{Rogerson, Richard, and Robert Shimer}
{2011}{Search in Macroeconomic Models of the Labor Market}
{Orley Ashenfelter and David Card}{Handbook of Labor Economics.
{\rm Volume 4A}}{Amsterdam: Elsevier, pp. 619--700}

 \article{Rogerson, Richard, and Johanna Wallenius}{2009}{Micro
and Macro Elasticities in a Life Cycle Model with Taxes.}{Journal of
Economic Theory}{Vol. 144, pp. 2277--2292}



 \ocontribution{Rogoff, Kenneth}{1989}{Reputation, Coordination, and Monetary
 Policy}{Robert J. Barro}{Modern Business Cycle Theory.}{Cambridge, Mass.:
 Harvard University Press, pp. 236--264}


 \book{Roll, Richard}{1970}{The Behavior of Interest Rates: An Application of
 the Efficient Market Model to U.S. Treasury Bills}{New York: Basic Books}


 \book{Romer, David}{1996}{Advanced Macroeconomics}{New York: McGraw-Hill}


\article{Romer, Paul M}{1986}{Increasing Returns and Long-Run Growth.}
{Journal of Political Economy}{Vol. 94, pp. 1002--1037}


\article{Romer, Paul M}{1987}{Growth Based on Increasing Returns Due to Specialization.}{American Economic Review Paper and Proceedings}
{Vol. 77, pp. 56--62}


\article{Romer, Paul M}{1990}{Endogenous Technological Change.}
{Journal of Political Economy}{Vol. 98, pp. S71--S102}




\article{Rosen, Sherwin, and Robert H. Topel}{1988}{Housing Investment
in the United States.}{Journal of Political Economy}{Vol. 96(4),
pp. 718--740}


 \article{Rosen, Sherwin, Kevin M. Murphy, and Jose A. Scheinkman}{1994}{Cattle
 Cycles.}{Journal of Political Economy}{Vol. 102(3), pp. 468-492}


 \article{Ross, Stephen A}{1976}{The Arbitrage Theory of Capital Asset
 Pricing.}{Journal of Economic Theory}{Vol. 13(3), pp. 341--360}


 \ocontribution{Rotemberg, Julio J}{1987}{The New Keynesian Microfoundations}
 {Stanley Fischer}{NBER Macroeconomics Annual.}
 {Cambridge, Mass.: MIT Press, pp. 69--104}




\contribution{Rotemberg, Julio J. and Michael Woodford}{1997}{An
Optimization-Based Econometric Framework for the Evaluation of
Monetary Policy}{Olivier Blanchard and Stanley Fischer}{NBER
Macroeconomic Annual}{Cambridge, Mass.: MIT Press, pp.
297-345}


 \article{Rothschild, Michael, and Joseph Stiglitz}{1970}{Increasing Risk I: A
 Definition.}{Journal of Economic Theory}{Vol. 2(3), pp. 225--243}


\article{Rothschild, Michael, and Joseph Stiglitz}{1971}{Increasing Risk II: Its Economic Consequences.}{Journal of Economic Theory}{Vol. 3(1), pp. 66--84}

\article{Rothschild, Michael}{1973}{Models of Market Organization with Imperfect Information: A Survey}{Journal of Political Economy}
{Vol. 81(6), pp. 1283--1308}



\article{Rubinstein, Mark}{1974}{An Aggregation Theorem for Security
Markets.}{Journal of Financial Economics}{Vol. 1(3), pp. 225--244}




\article{Runkle, David E}{1991}{Liquidity Constraints and the
Permanent-Income Hypothesis: Evidence from Panel Data.}{Journal of
Monetary Economics}{Vol. 27(1), pp.  73--98}

%
%\mimeo{Ryoo, Jaewoo, and Sherwin Rosen}{2003}{The Engineering
%Labor Market.}
%{Hong Kong University of Science and Technology}


\article{Ryoo, J. and S. Rosen}{2004}{The Engineering Labor Market.}
{Journal of Political Economy}{Vol.  112(1), pp. S110--S140}


 \article{Saint-Paul, Gilles}{1995}{The High Unemployment Trap.}
 {Quarterly Journal of Economics}{Vol. 110, pp. 527--550}


 \article{Samuelson, Paul A}{1958}{An Exact Consumption-Loan Model of
 Interest with or without the Social Contrivance of Money.}
 {Journal of Political Economy}{Vol. 66, pp. 467--482}


 \article{Samuelson, Paul A}{1965}{Proof That Properly Anticipated Prices
 Fluctuate Randomly.}{Industrial Management Review}{Vol. 6(1), pp. 41--49}




\article{Sandmo, Agnar}{1970}{The Effect of Uncertainty on Saving
Decisions.}{Review of Economic Studies}{Vol. 37, pp. 353--360}

\article{Santos, Manuel S}{1991}{Smoothness of the Policy Function in Discrete Time Economic Models.}{Econometrica}{Vol. 59(5), pp. 1365-82}

\article{Santos, Manuel S}{1993}{On High-Order Differentiability of the Policy Function}{Economic Theory}{Vol. 3(3), pp. 565-70}





\book{Sargent, Thomas J}{1979a}{Macroeconomic Theory}{New York:
Academic Press}

\contribution{Sargent, Thomas J}{1980}{Tobin's $q$ and
 the Rate of Investment in
 General Equilibrium}{K. Brunner and A. Meltzer}{On the State of
 Macroeconomics}{Carnegie-Rochester Conference Series 12, Amsterdam: North-Holland,
 pp. 107--154}

\mimeo{Sargent, Thomas J}{1980}{Lecture notes on Filtering, Control, and
 Rational Expectations.}{Minneapolis: University of Minnesota, Department of
 Economics}


\article{Sargent, Thomas J.}{1979b}{A Note On Maximum Likelihood
Estimation of The Rational Expectations Model of The Term
Structure}{Journal Of Monetary Economics}{Vol. 35, pp. 245--274}


 \book{Sargent, Thomas J}{1987a}{Macroeconomic Theory,}{2nd ed.
 New York: Academic Press}


 \book{Sargent, Thomas J}{1987b}{Dynamic Macroeconomic Theory}
 {Cambridge, Mass.: Harvard University Press}





 \article{Sargent, Thomas J}{1991}{Equilibrium with Signal Extraction
 from Endogeneous Variables.}{Journal of Economic Dynamics and Control}
 {Vol. 15, pp. 245--273}


\book{Sargent, Thomas J}{2013}{Rational Expectations and Inflation, 3rd ed.}
{Princeton: Princeton University Press}

\book{Sargent, Thomas J}{1999}{The Conquest of American Inflation}{Princeton University
Press, Princeton, New Jersey}

\article{Sargent, Thomas J}{2008}{Evolution and Intelligent Design}{American Economic Review}
 {Vol. 98(1), pp. 5--37}


\article{Sargent, Thomas J., and Bruce Smith}{1997}{Coinage, Debasements,
and Gresham's Laws.}{Economic Theory}{Vol. 10, pp. 197--226}


\article{Sargent, Thomas J., and Fran\c cois R. Velde}{1990}{The Analytics of
German Monetary Reform.}{Quarterly Review}{Federal Reserve
Bank of San Francisco, No. 4, pp. 33--50}


 \article{Sargent, Thomas J., and Fran\c cois R. Velde}{1995}{Macroeconomic
 Features of the French Revolution.}{Journal of Political Economy}
 {Vol. 103(3), pp. 474--518}


\article{Sargent, Thomas J., and Fran\c cois R. Velde}{1999}{The Big Problem of
Small Change.}{Journal of Money, Credit, and Banking}{Vol. 31(2), pp. 137--161}


 \article{Sargent, Thomas J., and Neil Wallace}{1973}{Rational Expectations
 and the Dynamics of Hyperinflation.}{International Economic Review}
 {Vol. 14, pp. 328--350}


 \article{Sargent, Thomas J., and Neil Wallace}{1975}{`Rational' Expectations,
the Optimal Monetary Instrument, and the Optimal Money Supply Rule}
{Journal of Political Economy}{Vol. 83(2), pp. 241--254}


 \article{Sargent, Thomas J., and Neil Wallace}{1981}{Some Unpleasant Monetarist
 Arithmetic.}{Quarterly Review}{Federal Reserve Bank of Minneapolis, Vol.
 5(3), pp. 1--17}


 \article{Sargent, Thomas J., and Neil Wallace}{1982}{The Real Bills Doctrine
 vs. the Quantity Theory: A Reconsideration.}{Journal of Political
 Economy}{Vol. 90(6), pp. 1212--1236}


 \article{Sargent, Thomas J., and Neil Wallace}{1983}{A Model of Commodity
 Money.}{Journal of Monetary Economics}{Vol. 12(1), pp. 163--187}


\article{Sargent, Thomas J., Noah Williams, and  Tao Zha}{2009}{The Conquest of South American Inflation.}
{Journal of Political Economy}{Vol. 117, No. 2, pp. 211--256}

\book{Scheinkman, Jose}{2014}{Speculation, Trading, and Bubbles}{Kenneth J. Arrow Lecture Series. New York:
Columbia University Press}


\article{Schmitt-Grohe, Stephanie, and Martin Uribe}{2004a}{Optimal
Fiscal and Monetary Policy Under Sticky Prices.} {Journal of
Economic Theory}{Vol. 114(2), pp. 198--230}

\article{Schmitt-Grohe, Stephanie and Martin Uribe}{2004b}{Solving
dynamic general equilibrium models using second-order
approximation to the policy function.}{Journal of Economic
Dynamics and Control}{Vol. 28, pp. 755-775}

\article{Seater, John J}{1993}{Ricardian Equivalence.}{Journal of
Economic Literature}{Vol. 31(1), pp. 142--190}




\article{Segerstrom, Paul S}{1998}{Endogenous Growth without Scale Effects.}
{American Economic Review}{Vol. 88, pp. 1290--1310}


\article{Segerstrom, Paul S., T. C. A. Anant, and Elias Dinopoulos}{1990}
{A Schumpeterian Model of the Product Life Cycle.}{American Economic Review}
{Vol. 80, pp. 1077--1091}


 \article{Shavell, Stephen, and Laurence Weiss}{1979}{The Optimal Payment of
 Unemployment Insurance Benefits Over Time.}{Journal of Political
 Economy}{Vol. 87, pp. 1347--1362}


\article{Shell, Karl}{1971}{Notes on the Economics of
Infinity.}{Journal of Political Economy}{Vol. 79(5), pp.
1002--1011}




\article{Shi, Shouyong}{1995}{Money and Prices: A Model of Search and
Bargaining.}{Journal of Economic Theory}{Vol. 67, pp. 467--496}


 \dissertation{Shiller, Robert J}{1972}{Rational Expectations and the Structure
 of Interest Rates}{Massachusetts Institute of Technology}


 \article{Shiller, Robert J}{1981}{Do Stock Prices Move Too Much to Be
 Justified by Subsequent Changes in Dividends?}{American Economic
 Review}{Vol. 71(3), pp. 421--436}



\article{Shimer, Robert}{2005}{The Cyclical Behavior of Equilibrium
Unemployment and Vacancies.}{American Economic Review}
{Vol. 95(1), pp. 25--49}

\book{Shimer, Robert}{2010}{Labor Markets and Business Cycles}
{Princeton, New Jersey: Princeton University Press}



\article{Shin, Yongseok}{2007}{Managing the maturity structure
of government debt.}{Journal of Monetary Economics}{Vol. 54, pp. 1565-1571}


\article{Sibley, David S.}{1975}{Permanent and Transitory Income Effects
in a Model of Optimal Consumption with Wage Income Uncertainty.}
{Journal of Economic Theory}{Vol. 11, pp. 68--82}




 \article{Sidrauski, Miguel}{1967}{Rational Choice and Patterns of Growth in a
 Monetary Economy.}{American Economic Review}{Vol. 57(2), pp. 534--544}


 \article{Sims, Christopher A}{1972}{Money, Income, and Causality.}{American
 Economic Review}{Vol. 62(4), pp. 540--552}


 \mimeo{Sims, Christopher A}{1989}{Solving Nonlinear Stochastic
 Optimization and Equilibrium Problems Backwards.}{Minneapolis: Institute for Empirical
 Macroeconomics, Federal Reserve Bank of Minneapolis, No. 15}


 \article{Sims, Christopher A}{1994}{A Simple Model for the Determination
 of the Price Level and the Interaction of Monetary and Fiscal Policy.}
 {Economic Theory}{Vol. 4, pp. 381--399}

\article {Siow, A}{1984} {Occupational Choice under Uncertainty.}
{Econometrica}{Vol. 52(3), pp. 631--645}



\article{Siu, Henry E}{2004}{Optimal Fiscal and Monetary Policy With Sticky
Prices.}{Journal of Monetary Economics}{Vol. 51, no. 3, pp. 575--607}



% \mimeo{Sims, Christopher A}{1989}{Solving Nonlinear Stochastic
% Optimization and Equilibrium Problems Backwards.}{Institute for Empirical
% Macroeconomics, Federal Reserve Bank of Minneapolis, 15}


\article{Smith, Bruce}{1988}{Legal Restrictions, `Sunspots,' and
Peel's Bank Act: The Real Bills Doctrine versus the Quantity Theory
 Reconsidered.}{Journal of Political Economy}{Vol. 96(1), pp. 3-19}


 \article{Smith, Lones}{1992}{Folk Theorems in Overlapping Generations Games.}
 {Games and Economic Behavior}{Vol. 4(3), pp. 426--449}


 \article{Solow, Robert M}{1956}{A Contribution to the Theory of Economic
 Growth.}{Quarterly Journal of Economics}{Vol. 70, pp. 65--94}


\article{Sotomayor, Marlida A. de Oliveira}{1984}
{On Income Fluctuations and Capital Gains.}
{Journal of Economic Theory}{Vol. 32(1), pp. 14--35}


\article{Spear, Stephen E., and Sanjay Srivastava}{1987}
{On Repeated Moral Hazard with Discounting.}
{Review of Economic Studies}{Vol. 54(4), pp. 599--617}


 \mimeo{Stacchetti, Ennio}{1991}{Notes on Reputational Models in
 Macroeconomics.}{Stanford University, September}


\book{Stachurski, John}{2009}{Economic Dynamics: Theory and Computation}
{Cambridge, Mass: MIT Press}

 \article{Stigler, George}{1961}{The Economics of Information.}
 {Journal of Political Economy}{Vol. 69(3), pp. 213--225}


 \article{Stiglitz, Joseph E}{1969}{A Reexamination of the Modigliani-Miller
 Theorem.}{American Economic Review}{Vol. 59(5), pp. 784--793}


\contribution{Stiglitz, Joseph E}{1987}{Pareto Efficient and Optimal Taxation
and the New New Welfare Economics}{Alan J. Auerbach, and Martin Feldstein}
{Handbook of Public Economics, Vol. 2}{Amsterdam: Elsevier/North-Holland}


 \article{Stokey, Nancy L}{1989}{Reputation and Time Consistency.}
 {American Economic Review}{Vol. 79, pp. 134--139}


 \article{Stokey, Nancy L}{1991}{Credible Public Policy.}{Journal of Economic
 Dynamics and Control}{Vol. 15(4), pp. 627--656}


 \book{Stokey, Nancy, and Robert E. Lucas, Jr. (with Edward C. Prescott)}{1989}
 {Recursive Methods in Economic Dynamics}{Cambridge, Mass.: Harvard University Press}


 \mimeo{Storesletten, Kjetil, Christopher I. Telmer, and Amir Yaron}{1998}
 {Persistent Idiosyncratic Shocks and Incomplete Markets.}{Carnegie Mellon
 University and Wharton School, University of Pennsylvania}

 \article{Storesletten, Kjetil, Christopher I. Telmer, and Amir Yaron}{2004}
{Consumption and risk sharing over the life cycle.}{Journal of Monetary Economics}{Vol. 51(3), pp. 609-633}



\mimeo{Straub, Ludwig, and Iv\'an Werning}{2015}{Positive Long Run
Capital Taxation: Chamley-Judd Revisited}{Massachusetts Institute
of Technology}

 \article{Svensson, Lars E. O}{1986}{Sticky Goods Prices, Flexible
 Asset Prices, Monopolistic Competition, and Monetary Policy.}
 {Review of Economic Studies}{Vol. 53, pp. 385--405}






 \dissertation{Tallarini, Thomas D., Jr}{1996}{Risk-Sensitive
 Real Business Cycles}{University of Chicago}


\article{Tallarini, Thomas D., Jr}{2000}{Risk-Sensitive Real Business
Cycles.}{Journal of Monetary Economics}{Vol. 45(3), pp. 507--532}


 \article{Tauchen, George}{1986}{Finite State Markov Chain Approximations
 to Univariate and Vector Autoregressions.}{Economic Letters}
 {Vol. 20, pp. 177--181}


 \article{Taylor, John B}{1977}{Conditions for Unique Solutions
 in Stochastic Macroeconomic Models with Rational Expectations.}
 {Econometrica}{Vol. 45, pp. 1377--1185}


 \article{Taylor, John B}{1980}{Output and Price Stability:
 An International Comparison.}{Journal of Economic Dynamics and Control}
 {Vol. 2, pp. 109--132}


 \article{Thomas, Jonathan, and Tim Worrall}{1988}{Self-Enforcing Wage
 Contracts.}{Review of Economic Studies}{Vol. 55, pp. 541--554}


 \article{Thomas, Jonathan, and Tim Worrall}{1990}{Income Fluctuation
 and Asymmetric Information: An Example of a Repeated Principal-Agent
 Problem.}{Journal of Economic Theory}{Vol. 51(2), pp. 367--390}


\mimeo{Thomas, Jonathan, and Tim Worrall}{1994}{Informal Insurance
Arrangements in Village Economies.}{Liverpool Research Papers in
Economics and Finance, 9402}

 \article{Tirole, Jean}{1982}{On the Possibility of Speculation under Rational
 Expectations.}{Econometrica}{Vol. 50, pp. 1163--1181}


 \article{Tirole, Jean}{1985}{Asset Bubbles and Overlapping Generations.}
 {Econometrica}{Vol. 53, pp. 1499--1528}


 \article{Tobin, James}{1956}{The Interest Elasticity of the Transactions Demand
 for Cash.}{Review of Economics and Statistics}{Vol. 38, pp. 241--247}


 \article{Tobin, James}{1961}{Money, Capital, and Other Stores of Value}
 {American Economic Review}{Vol. 51(2), pp. 26--37}


 \pcontribution{Tobin, James}{1963}{An Essay on the Principles of Debt
 Management}{William Fellner et al.}{Fiscal and Debt Management
 Policies.}{Englewood Cliffs, N.J.: Prentice-Hall, pp. 141--215.
 (Reprinted in James Tobin, {\it Essays in Economics}, Vol. 1.
 Amsterdam: North-Holland, 1971, pp. 378--455.)}


 \article{Topel, Robert H., and Sherwin Rosen}{1988}{Housing Investment
 in the United States.}{Journal of Political Economy}
 {Vol. 96(4), pp. 718--740}


\contribution{Townsend, Robert M}{1980}{Models of Money with Spatially Separated Agents}
{J. H. Kareken and N. Wallace}{Models of Monetary Economies}
{Minneapolis: Federal Reserve Bank of Minneapolis, pp. 265--303}


 \article{Townsend, Robert M}{1983}{Forecasting the Forecasts of Others.}
 {Journal of Political Economy}{Vol. 91, pp. 546--588}




\article{Townsend, Robert M}{1994}{Risk and Insurance in Village
India}{Econometrica}{Vol. 62}{539--592}.






 \article{Trejos, Alberto, and
Randall Wright}{1995}{Search, Bargaining, Money and
Prices.}{Journal of Political Economy}{Vol. 103, pp. 118--139}


 \article{Turnovsky, Stephen J., and William A. Brock}{1980}{Time Consistency
 and Optimal Government Policies in Perfect Foresight Equilibrium.}
 {Journal of Public Economics}{Vol. 13, pp. 183--212}






\article{Uzawa, Hirofumi}{1965}{Optimum Technical Change in an Aggregative Model
of Economic Growth.}{International Economic Review}{Vol. 6, pp. 18--31}






\article{Villamil, Anne P}{1988}{Price Discriminating Monetary Policy:
A Nonuniform Pricing Approach.}{Journal of Public Economics}
{Vol. 35(3), pp. 385--392}






\contribution{Wallace, Neil}{1980}{The Overlapping Generations Model of Fiat Money}{J. H. Kareken and N. Wallace}{Models of Monetary Economies}{Minneapolis: Federal Reserve Bank of Minneapolis, pp. 49--82}


 \article{Wallace, Neil}{1981}{A Modigliani-Miller Theorem for Open-Market
 Operations.}{American Economic Review}{Vol. 71, pp. 267--274}


 \article{Wallace, Neil}{1983}{A Legal Restrictions Theory of the Demand
 for `Money' and the Role of Monetary Policy.}{Quarterly Review}
 {Federal Reserve Bank of Minneapolis, Vol. 7(1), pp. 1--7}


 \ocontribution{Wallace, Neil}{1989}{Some Alternative Monetary Models and
 Their Implications for the Role of Open-Market Policy}{Robert J. Barro}
 {Modern Business Cycle Theory.}{Cambridge, Mass.: Harvard University Press, pp. 306--328}


\book{Walsh, Carl E}{1998}{Monetary Theory and Policy}{Cambridge, Mass.:
MIT Press}




 \article{Wang, Cheng, and Stephen D. Williamson}{1996}{Unemployment Insurance
 with Moral Hazard in a Dynamic Economy.}{Carnegie-Rochester Conference
 Series on Public Policy}{Vol. 44, pp. 1--41}

\article{Wang, Neng}{2003}{Caballero meets Bewley: The permanent-income hypothesis in general equilbrium}
{American Economic Review}{Vol. 93(3), pp. 927--936}


 \book{Watanabe, Shinichi}{1984}{Search Unemployment, the Business Cycle, and
 Stochastic Growth}{Ph.D. dissertation, University of Minnesota}


 \article{Weil, Philippe}{1989}{The Equity Premium Puzzle and the Risk-Free Rate
 Puzzle.}{Journal of Monetary Economics}{Vol. 24(2), pp. 401--421}


\article{Weil, Philippe}{1990}{Nonexpected Utility in Macroeconomics.}
{Quarterly Journal of Economics}{Vol. 105, pp. 29--42}


 \article{Weil, Philippe}{1993}{Precautionary Savings and the Permanent Income
 Hypothesis.}{Review of Economic Studies}{Vol. 60(2), pp. 367--383}


 \mimeo{Werning, Ivan}{2002}{Optimal unemployment insurance with
 hidden savings.}{University of Chicago Working Paper}

\article{Werning, Ivan}{2007}{Optimal Fiscal Policy with Redistribution.}
{Quarterly Journal of Economics}{Vol. 122(3),  pp. 925--67}


 \article{Wen, Y}{1998}{Can a Real Business Cycle Pass the Watson
 Test?}{Journal of Monetary Economics}{Vol. 42, pp. 185--202}


 \book{Whiteman, Charles H}{1983}{Linear Rational Expectations Models: A Users
 Guide}{Minneapolis: University of Minnesota Press}


 \book{Whittle, Peter}{1963}{Prediction and Regulation by Linear
 Least-Square Methods}{Princeton, N.J.: Van Nostrand-Reinhold}


\book{Whittle, Peter}{1990}{Risk-Sensitive Optimal Control}{New York: Wiley}


 \article{Wilcox, David W}{1989}{The Sustainability of Government Deficits:
 Implications of the Present-Value Borrowing Constraint.}{Journal of Money,
 Credit, and Banking}{Vol. 21(3), pp. 291--306}


 \article{Woodford, Michael}{1994}{Monetary Policy and Price Level Determinacy
 in a Cash-in-Advance Economy.}{Economic Theory}{Vol. 4, pp. 345--380}


 \article{Woodford, Michael}{1995}{Price-Level Determinacy without Control
 of a Monetary Aggregate.}{Carnegie-Rochester Conference Series
 on Public Policy}{Vol. 43, pp. 1--46}


 \article{Woodford, Michael}{2003}{Optimal Interest-Rate
 Smoothing.}
{Review of Economic Studies}{Vol. 70(4), pp.  861--886}




\book{Woodford, Michael}{2003}{Interest and Prices}{Princeton,
N.J.: Princeton University Press}


 \article{Wright, Randall}{1986}{Job Search and Cyclical Unemployment.}
 {Journal of Political Economy}{Vol. 94(1), pp. 38--55}






\article{Young, Alwyn}{1998}{Growth without Scale Effects.}
{Journal of Political Economy}{Vol. 106, pp. 41--63}


\article{Zeira, Joseph}{1999}{Informational Overshooting, Booms, and
Crashes.}{Journal of Monetary Economics}{Vol. 43(1), pp. 237--257}


 \article{Zeldes, Stephen P}{1989}{Optimal Consumption with Stochastic Income:
 Deviations from Certainty Equivalence.}{Quarterly Journal of Economics}{Vol.
 104(2), pp. 275--298}


\article{Zhang, Harold}{1997}{Endogenous Borrowing Constraints
with Incomplete Markets.} {Journal of Finance}{Vol. 52(5), pp.
2187--2209}






\mimeo{Zhao, Rui}{2001}{The Optimal Unemployment Insurance Contract:
Why a Replacement Ratio?}{University of Illinois, Champagne-Urbana, Department of Economics}

\mimeo{Zhu, Shenghao}{2009}{Existence of Equilibrium in an Incomplete Market Model with Endogenous Labor Supply}
{New York University}


\article{Zhu, Xiaodong}{1992}{Optimal Fiscal Policy in a Stochastic Growth
Model.}{Journal of Economic Theory}{Vol. 58, pp. 250--289}

\article{Zin, Stanley E}{2002}{Are behavioral asset-pricing models structural?}{Journal of Monetary Economics}
{Vol. 49(1), pp. 215-228}



}% end eightpoint
